---
title: "Pre-process the data for PENCIL input"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r}
# clear all objects including hidden objects
rm(list = ls(all.names = TRUE)) 
# free up memory and report the memory usage
gc()
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
require(ggplot2)
ggplot2::theme_set(new = theme_bw(base_size = 12)) # sets the theme and axis font size
```



# Data description:

1. GSE200996

Original paper: Luoma, Adrienne M., et al. "Tissue-resident memory and circulating T cells are early responders to pre-surgical cancer immunotherapy." Cell 185.16 (2022): 2918-2935. https://doi.org/10.1016/j.cell.2022.06.018

Data type: matched tumor and blood, scRNA-seq, scTCR-seq, bulk TCR; Samples: from 30 patients from 2 different cohorts (anti-PD-1 monotherapy, anti-PD-1+anti-CTLA-4 combination therapy)


# Data input:

Data matrix downloaded from GEO

# Sample description (no healthy donors)

Blood B1 (pre): P01, P02, P04, P05, P08, P09, P10, P12, P13, P14, P15, P16, P17, P18, P19, P20, P21, P22, P23, P24, P25, P26, P28, P29, P30, P31, P32

Blood B2 (post): P01, P02, P04, P05, P08, P09, P10, P12, P13, P14, P15, P16, P17, P19, P20, P22, P23, P24, P26, P28, P29, P30, P31, P32

Blood B3 (follow-up): P01, P02, P04, P05, P09, P10, P12, P13, P14, P15, P17, P19, P20, P21, P22, P23, P25, P32

Tumor pre: P18, P23, P24, P29, P32, [P27 (do not have blood match)]

Tumor post: P13, P14, P15, P16, P18, P19, P20, P21, P22, P23, P24, P25, P26, P27, P28, P29, P30, P31, P32



# Aims:

- integrate all patients' data object

- analysis the integrated data

- annotate the data with celltypist

- save the integrated seurat object

# Connect to server
```{bash}
ssh changt7@biowulf.nih.gov
sinteractive --mem=64g --cpus-per-task=32 --time=24:00:00
sinteractive --mem=128g --cpus-per-task=32 --gres=gpu:k80:2 --time=24:00:00
# --gres=lscratch:1024
module load R/
R

```

# load required package
```{r, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("UCell")
library(UCell)

#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)

library(caret)
library(Seurat)
library(GEOquery)
library(dplyr)
library(ggplot2)
library(SingleR)
library(sctransform)
library(celldex)
library(dittoSeq)
library(SeuratWrappers)
library(harmony)
library(scuttle)
library(scran)
library(scFeatureFilter)
library(cowplot)
library(patchwork)
library(edgeR)
library(RColorBrewer)
library(stringr)

```

# set input and output directories & parameters
```{r}

dataset = 'GSE200996'

# local
data_dir <- paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/05.PENCIL/02.Input/", dataset, '/')
processed_dir <- paste0("../03.Results/", dataset, '/')
fig_dir <- paste0("../03.Results/", dataset, '/Figures/')

# server
shared_data_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/02.Input/")
data_dir <- paste0(shared_data_dir, dataset, '/')
processed_dir <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset, '/')
fig_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset, "/Figures/")
PENCIL_result_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/01.Scripts/results/")
```

# Add phenotype info for GSE200996 (Luoma et al., HNSCC, ICB dataset)
```{r}

# Read in phenotype info.
phenotype_df = read.csv(file = paste0(data_dir,"clinical_info.csv"))
phenotype_df = phenotype_df[,c(1,7,9,10,15:26)]
colnames(phenotype_df) = c(colnames(phenotype_df)[1:4],'Downstaging','RECIST_response','Volumetric_response','Viable_Tumor', 'Pathological_response', 'Pathological_response_binned', 'Max_response', 'Any_response', 'OS_days', 'PFS_days', 'PFS_event', 'OS_event')

# Add phenotype info to seurat object
meta <- merged_seu@meta.data
meta$cell_names <- rownames(meta) # save row names before merge with phenotype data
meta.new <- merge(meta, phenotype_df, by.x = "patient", by.y = "patient_id")
rownames(meta.new) <- meta.new$cell_names # retrieve rownames 
merged_seu <- AddMetaData(merged_seu, meta.new)
#cell.names <- sapply(seq_along(rownames(merged_seu@meta.data)), function(i) paste0("cell_", i), USE.NAMES = F)
#merged_seu@meta.data$cellID <- cell.names # rename cells to cell_1, cell_2, ...
#merged_seu <- RenameCells(object = merged_seu, new.names = merged_seu@meta.data$cellID)
#print(merged_seu$cellID[1:10])

saveRDS(object = merged_seu, file = paste0(data_dir, "seu_",sampleType_name,"_all_raw.rds")) 



```


