---
title: "Pre-process the data for PENCIL input"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r}
# clear all objects including hidden objects
rm(list = ls(all.names = TRUE)) 
# free up memory and report the memory usage
gc()
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
require(ggplot2)
ggplot2::theme_set(new = theme_bw(base_size = 12)) # sets the theme and axis font size
```



# Data description:

1. GSE200996

Original paper: Luoma, Adrienne M., et al. "Tissue-resident memory and circulating T cells are early responders to pre-surgical cancer immunotherapy." Cell 185.16 (2022): 2918-2935. https://doi.org/10.1016/j.cell.2022.06.018

Data type: matched tumor and blood, scRNA-seq, scTCR-seq, bulk TCR; Samples: from 30 patients from 2 different cohorts (anti-PD-1 monotherapy, anti-PD-1+anti-CTLA-4 combination therapy)


# Data input:

Data matrix downloaded from GEO

# Sample description (no healthy donors)

Blood B1 (pre): P01, P02, P04, P05, P08, P09, P10, P12, P13, P14, P15, P16, P17, P18, P19, P20, P21, P22, P23, P24, P25, P26, P28, P29, P30, P31, P32

Blood B2 (post): P01, P02, P04, P05, P08, P09, P10, P12, P13, P14, P15, P16, P17, P19, P20, P22, P23, P24, P26, P28, P29, P30, P31, P32

Blood B3 (follow-up): P01, P02, P04, P05, P09, P10, P12, P13, P14, P15, P17, P19, P20, P21, P22, P23, P25, P32

Tumor pre: P18, P23, P24, P29, P32, [P27 (do not have blood match)]

Tumor post: P13, P14, P15, P16, P18, P19, P20, P21, P22, P23, P24, P25, P26, P27, P28, P29, P30, P31, P32



# Aims:

- integrate all patients' data object

- analysis the integrated data

- annotate the data with celltypist

- save the integrated seurat object

# load required package
```{r, include=FALSE}

library(Seurat)
library(GEOquery)
library(dplyr)
library(ggplot2)
library(SingleR)
library(sctransform)
library(celldex)
library(dittoSeq)
library(SeuratWrappers)
library(harmony)
library(edgeR)
library(scuttle)
library(scran)
library(scFeatureFilter)
library(cowplot)
library(patchwork)
library(edgeR)
library(RColorBrewer)
library(stringr)

```

# set input and output directories & parameters
```{r}

dataset = 'GSE169246' # 'GSE200996'   'GSE120575'   'GSE169246'

# local
data_dir <- paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/05.PENCIL/02.Input/", dataset, '/')
processed_dir <- paste0("../03.Results/", dataset, '/')
fig_dir <- paste0("../03.Results/", dataset, '/Figures/')

# server
data_dir <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/02.Input/", dataset, '/')
processed_dir <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset, '/')
fig_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset, "/Figures/")

```

# Connect to server
```{bash}
ssh changt7@biowulf.nih.gov
sinteractive --mem=64g --cpus-per-task=32 --time=24:00:00
sinteractive --mem=128g --cpus-per-task=32 --gres=gpu:k80:2 --time=24:00:00
# --gres=lscratch:1024
module load R/
R

```

# load seurat object
```{r}

sampleType_name = 'Tissue' # 'PBMC'  'Tissue'

merged_seu = readRDS(file = paste0(data_dir,"seu_",sampleType_name,".rds"))

# Normalize the data
#merged_seu <- NormalizeData(merged_seu) # log(x/sum(x)*10000+1)

# Find highly variable genes
merged_seu <- FindVariableFeatures(merged_seu)
# Scale the data
merged_seu <- ScaleData(merged_seu) # pbmc <- SCTransform(pbmc, vars.to.regress = "CC.Difference")



# classification mode
#exp_data = merged_seu@assays[["RNA"]]@scale.data[VariableFeatures(merged_seu),]
# regression mode
#exp_data = as.matrix(merged_seu@assays[["RNA"]]@counts)

```


# Add phenotype info for GSE200996 (Luoma et al., HNSCC, ICB dataset)
```{r}
#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)


# Read in phenotype info.
phenotype_df = read.csv(file = paste0(data_dir,"clinical_info.csv"))
phenotype_df = phenotype_df[,c(1,7,9,10,15:26)]
colnames(phenotype_df) = c(colnames(phenotype_df)[1:4],'Downstaging','RECIST_response','Volumetric_response','Viable_Tumor', 'Pathological_response', 'Pathological_response_binned', 'Max_response', 'Any_response', 'OS_days', 'PFS_days', 'PFS_event', 'OS_event')

# Add phenotype info to seurat object
meta <- merged_seu@meta.data
meta$cell_names <- rownames(meta) # save row names before merge with phenotype data
meta.new <- merge(meta, phenotype_df, by.x = "patient", by.y = "patient_id")
rownames(meta.new) <- meta.new$cell_names # retrieve rownames 
merged_seu <- AddMetaData(merged_seu, meta.new)
#cell.names <- sapply(seq_along(rownames(merged_seu@meta.data)), function(i) paste0("cell_", i), USE.NAMES = F)
#merged_seu@meta.data$cellID <- cell.names # rename cells to cell_1, cell_2, ...
#merged_seu <- RenameCells(object = merged_seu, new.names = merged_seu@meta.data$cellID)
#print(merged_seu$cellID[1:10])

saveRDS(object = merged_seu, file = paste0(data_dir, "seu_",sampleType_name,".rds")) 



```


# Add phenotype info for GSE169246 (Zeming Zhang, TNBC, ICB dataset)
```{r}
#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)


# Read in phenotype info.
phenotype_df = read.csv(file = paste0(data_dir,"clinical_info.csv"))
phenotype_df = phenotype_df[,c(1,7,9,10,15:26)]
colnames(phenotype_df) = c(colnames(phenotype_df)[1:4],'Downstaging','RECIST_response','Volumetric_response','Viable_Tumor', 'Pathological_response', 'Pathological_response_binned', 'Max_response', 'Any_response', 'OS_days', 'PFS_days', 'PFS_event', 'OS_event')

# Add phenotype info to seurat object
meta <- merged_seu@meta.data
meta$cell_names <- rownames(meta) # save row names before merge with phenotype data
meta.new <- merge(meta, phenotype_df, by.x = "patient", by.y = "patient_id")
rownames(meta.new) <- meta.new$cell_names # retrieve rownames 
merged_seu <- AddMetaData(merged_seu, meta.new)
#cell.names <- sapply(seq_along(rownames(merged_seu@meta.data)), function(i) paste0("cell_", i), USE.NAMES = F)
#merged_seu@meta.data$cellID <- cell.names # rename cells to cell_1, cell_2, ...
#merged_seu <- RenameCells(object = merged_seu, new.names = merged_seu@meta.data$cellID)
#print(merged_seu$cellID[1:10])

saveRDS(object = merged_seu, file = paste0(data_dir, "seu_",sampleType_name,".rds")) 



```

# load raw expression and phenotype data and construct seurat object for GSE120575 (Sade-Feldman et al., melanoma, ICB dataset)
```{r}

library(readr)

sampleType_name = 'Tissue'

# read in log2(TPM+1) values (transcript-per-million reads) for 55,737 transcripts (rows) and 16,291 cells (columns)
data <- read.table(paste0(data_dir,"GSE120575_Sade_Feldman_melanoma_single_cells_TPM_GEO.txt"), header = TRUE, sep = "\t", row.names = 1, stringsAsFactors = FALSE, fill = TRUE) 
new_colnames = c(colnames(data)[2:16292],'X')
colnames(data) = new_colnames
data <- data[, -ncol(data)] # remove last NA column
patient_df <- data.frame(patient = t(data[1,]))
colnames(patient_df) = c('patient')
patient_df$patient <- sub("Post_P17_myeloid_enriched", "Post_P17", patient_df$patient)
patient_df$patient <- sub("Post_P19_myeloid_enriched", "Post_P19", patient_df$patient)
patient_df$patient <- sub("Post_P2_T_enriched", "Post_P2", patient_df$patient)
patient_df$patient <- sub("Post_P4_T_enriched", "Post_P4", patient_df$patient)
patient_df$patient <- sub("Post_P10_T_enriched", "Post_P10", patient_df$patient)
patient_df$patient <- sub("Post_P13_T_enriched", "Post_P13", patient_df$patient)
patient_df$patient <- sub("Post_P8_T_enriched", "Post_P8", patient_df$patient)
patient_df$patient <- sub("Post_P16_T_enriched", "Post_P16", patient_df$patient)
patient_df$patient <- sub("Post_P6_T_enriched", "Post_P6", patient_df$patient)
patient_df$cellID = rownames(patient_df)
data <- data[-1,] # remove patientID row



# Create a Seurat object
seurat_obj <- CreateSeuratObject(counts = data)
# Assign the normalized counts to seurat_obj[["RNA"]]@data
seurat_obj[["RNA"]]@data <- as.matrix(seurat_obj[["RNA"]]@counts)
# convert from log2(TPM+1) --> TPM
seurat_obj[["RNA"]]@data <- 2^seurat_obj[["RNA"]]@data-1 
# Create a new Seurat object with the TPM as counts
seurat_obj_new <- CreateSeuratObject(counts = seurat_obj[["RNA"]]@data)
seurat_obj = seurat_obj_new

# Assign project name
seurat_obj$project <- "GSE120575_Sade_Feldman_melanoma"

# add phenotype info
phenotype_df_raw <- read.table(paste0(data_dir,"GSE120575_singleCells_metaInfo_clean.txt"), sep = "\t", quote = "", header = TRUE, stringsAsFactors = FALSE)
phenotype_df = phenotype_df_raw[,c(5,6,7)] # "CellID" "PatientID" "ResponseInfo" "DrugType"  
colnames(phenotype_df) = c("patient", "ResponseInfo", "DrugType")
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
# keep only non-duplicated patientID --> dim(phenotype_df.new)=c(48,3)
phenotype_df.new <- merge(patient_df, phenotype_df, by.x = "patient", by.y = "patient")
rownames(phenotype_df.new) <- phenotype_df.new$cellID


# Add phenotype info to seurat object
# seurat_obj@meta.data$cellID <- NULL
# seurat_obj@meta.data$patient <- NULL
# seurat_obj@meta.data$ResponseInfo <- NULL
# seurat_obj@meta.data$DrugType <- NULL

meta <- seurat_obj@meta.data
meta$cell_names <- rownames(meta) # save row names before merge with phenotype data
meta.new <- merge(meta, phenotype_df.new, by.x = "cell_names", by.y = "cellID")
rownames(meta.new) <- meta.new$cell_names # retrieve rownames 
meta.new <- meta.new[, !(names(meta.new) %in% c("cellID", "cell_names"))]
seurat_obj <- AddMetaData(seurat_obj, meta.new)


#################################################################################
############# Cell annotation using 20.Cell_Annotation.Rmd pipeline #############
#################################################################################

##### Re-do CD8T cell annotation using original annotation by Sade-Feldman et al. 2018 Cell 

CD8T_cellID = read.csv(file = paste0(data_dir,"CD8T_original_cellAnno.csv"))

CD8T_cellID$Cell.Name <- gsub("-", ".", CD8T_cellID$Cell.Name)

merged_seu@meta.data$cell_type_original <- "Other"

merged_seu@meta.data$cell_type_original[rownames(merged_seu@meta.data) %in% CD8T_cellID$Cell.Name] <- "CD8+ T-cells"

#### compare self annotation and original annotation with UMAP
# Plot cell_type_original clusters
DimPlot(merged_seu, group.by = "cell_type_original", label = TRUE, repel = TRUE, raster=FALSE)+NoLegend()+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_cellTypeOriginalPaper_CD8T_",sampleType_name,".png"), height = 120, width = 140, dpi = 800, units = "mm")

# Plot cell_type CD8T clusters
merged_seu$highlighted.clusters <- ifelse(merged_seu$cell_type %in% c("CD8+ T-cells"), as.character(merged_seu$cell_type), "Other")
DimPlot(merged_seu, group.by = "highlighted.clusters", label = TRUE, repel = TRUE, raster=FALSE)+NoLegend()+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_cellTypeSelf_CD8T_",sampleType_name,".png"), height = 120, width = 140, dpi = 800, units = "mm")

```


# subset seurat to focus on cell types of interest (e.g. CD8T, CD4T, B, ...)
```{r}
############## CD8T cells
if (dataset == "GSE120575"){
  merged_seu_CD8Tcell <- subset(merged_seu, subset = cell_type_original %in% c("CD8+ T-cells"))
  # save to .h5ad for PENCIL to read
  SaveH5Seurat(merged_seu_CD8Tcell, filename = paste0(data_dir, "seu_",sampleType_name,"_CD8Tcell.h5Seurat"), overwrite = TRUE)
  Convert(paste0(data_dir, "seu_",sampleType_name,"_CD8Tcell.h5Seurat"), dest = "h5ad", assay = "RNA", overwrite = TRUE) # .X: only scaled data for HVGs. .raw.X: all genes log1p data.
  # saveRDS(object = merged_seu_CD8Tcell, file = paste0(data_dir, "seu_",sampleType_name,"_CD8Tcell.rds")) 
}else{
  merged_seu_CD8Tcell <- subset(merged_seu, subset = cell_type %in% c("CD8+ T-cells"))
  # save to .h5ad for PENCIL to read
  SaveH5Seurat(merged_seu_CD8Tcell, filename = paste0(data_dir, "seu_",sampleType_name,"_CD8Tcell.h5Seurat"), overwrite = TRUE)
  Convert(paste0(data_dir, "seu_",sampleType_name,"_CD8Tcell.h5Seurat"), dest = "h5ad", assay = "RNA", overwrite = TRUE) # .X: only scaled data for HVGs. .raw.X: all genes log1p data.
  # saveRDS(object = merged_seu_CD8Tcell, file = paste0(data_dir, "seu_",sampleType_name,"_CD8Tcell.rds")) 
}

############## CD4T cells
merged_seu_CD4Tcell <- subset(merged_seu, subset = cell_type %in% c("CD4+ T-cells", "Tregs"))
# save to .h5ad for PENCIL to read
SaveH5Seurat(merged_seu_CD4Tcell, filename = paste0(data_dir, "seu_",sampleType_name,"_CD4Tcell.h5Seurat"), overwrite = TRUE)
Convert(paste0(data_dir, "seu_",sampleType_name,"_CD4Tcell.h5Seurat"), dest = "h5ad", assay = "RNA", overwrite = TRUE) # Seems that only HVGs are saved to file. Other genes are left out.
# saveRDS(object = merged_seu_CD4Tcell, file = paste0(data_dir, "seu_",sampleType_name,"_CD4Tcell.rds")) 



############## T cells
merged_seu_Tcell <- subset(merged_seu, subset = cell_type %in% c("CD4+ T-cells", "CD8+ T-cells", "Tregs"))
# save to .h5ad for PENCIL to read
SaveH5Seurat(merged_seu_Tcell, filename = paste0(data_dir, "seu_",sampleType_name,"_Tcell.h5Seurat"), overwrite = TRUE)
Convert(paste0(data_dir, "seu_",sampleType_name,"_Tcell.h5Seurat"), dest = "h5ad", assay = "RNA", overwrite = TRUE) # Seems that only HVGs are saved to file. Other genes are left out.
#saveRDS(object = merged_seu_Tcell, file = paste0(data_dir, "seu_",sampleType_name,"_Tcell.rds")) 


############## B cells
merged_seu_Bcell <- subset(merged_seu, subset = cell_type %in% c("Naive B-cells", "Memory B-cells", "Plasma cells"))
# save to .h5ad for PENCIL to read
SaveH5Seurat(merged_seu_Bcell, filename = paste0(data_dir, "seu_",sampleType_name,"_Bcell.h5Seurat"), overwrite = TRUE)
Convert(paste0(data_dir, "seu_",sampleType_name,"_Bcell.h5Seurat"), dest = "h5ad", assay = "RNA", overwrite = TRUE) # Seems that only HVGs are saved to file. Other genes are left out.
#saveRDS(object = merged_seu_Bcell, file = paste0(data_dir, "seu_",sampleType_name,"_Bcell.rds")) 


#print(unique(merged_seu$patient))
#print(merged_seu$RECIST_response[1:10])
#print(merged_seu$Any_response[1:10])
#print(unique(merged_seu$timePt))
#print(dim(merged_seu))
```


# Run PENCIL
```{r}

```




# Analyze PENCIL identified phenotypic cells for tumor samples
```{r}

predicted_cell_labels = read.csv("/data/Lab_ruppin/tiangen/CancerProject/05.scRNAseq_project/03.Results/PENCIL/GSE200996/py/Volumetric_response/predicted_labels.csv") # Blood_Volumetric_response   Volumetric_response
predicted_cell_labels_new <- subset(predicted_cell_labels, select = 'predicted_label')
meta <- tumor_CD45_cells@meta.data
meta.new <- cbind(meta, predicted_cell_labels_new) # cell ID in PENCIL output does not match the original ones. But it is in the same order. So here we directly cbind two data frames to add the predicted label information for all cells.
tumor_CD45_cells <- AddMetaData(tumor_CD45_cells, meta.new)

```

# Plot predicted labels to confirm 
```{r}
DimPlot(tumor_CD45_cells, group.by = "predicted_label")+
  # scale_color_manual(values = my.cols)+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_predicted_label.png"), height = 120, width = 190, dpi = 800, units = "mm")
```

# Calculate R/NR cell number and ratios in each patient
```{r}
# Calculate the cell number for each patient
grouped <- meta.new %>%
  group_by(patient, predicted_label)

# Calculate the number of cells for each patient and predicted_label
cell_counts <- grouped %>%
  summarise(cell_count = n())
cell_counts <- cell_counts[cell_counts$predicted_label != "Rejected", ]

# Create a vector with all unique patient IDs
all_patients <- unique(cell_counts$patient)
# Create a new data frame with all patients having both 'yes' and 'no'
new_df <- data.frame(patient = rep(all_patients, each = 2),
                     predicted_label = rep(c("yes", "no"), length(all_patients)))
# Merge the new data frame with the original data frame based on patient ID and 'yes_no'
cell_counts_new <- merge(cell_counts, new_df, by = c("patient", "predicted_label"), all = TRUE)
# Replace missing values (NA) in the 'values' column with 0
cell_counts_new$cell_count[is.na(cell_counts_new$cell_count)] <- 0


# Calculate the total count for each patient
df_totals <- aggregate(cell_count ~ patient, cell_counts_new, sum)
# Calculate the count ratios for 'yes' and 'no'
df_ratios <- aggregate(cell_count ~ patient + predicted_label, cell_counts_new, sum)
df_ratios <- transform(df_ratios, ratio = cell_count / df_totals$cell_count * 100)

# Plot the stacked bar plot
ggplot(df_ratios, aes(x = patient, y = ratio, fill = predicted_label)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("yes" = "blue", "no" = "red")) +
  labs(x = "Patient", y = "Count Ratio", fill = "Response") +
  ggtitle("Count Ratios of 'R' and 'NR' Cells for Each Patient") +
  theme_minimal()
ggsave(filename = paste0(fig_dir, "R_NR_cellRatio_stackedBar.png"), height = 120, width = 190, dpi = 800, units = "mm")
```

# Cell annotation for R and NR cells in tumor
```{r}

tumor_CD45_cells_R_NR = tumor_CD45_cells[,grepl("yes", tumor_CD45_cells$predicted_label) | grepl("no", tumor_CD45_cells$predicted_label)]

ref <- celldex::DatabaseImmuneCellExpressionData() # celldex::BlueprintEncodeData()
singleR <- as.SingleCellExperiment(tumor_CD45_cells_R_NR, assay= "RNA") 
singler.anno <- SingleR(test = singleR, ref = ref, labels = (ref$label.fine))
tumor_CD45_cells_R_NR[["DICE.fine"]] <- singler.anno$labels # Database of Immune Cell Expression (DICE).
tumor_CD45_cells_R_NR[["DICE.fine.pruned"]] <- singler.anno$pruned.labels
saveRDS(tumor_CD45_cells_R_NR, file = "/data/Lab_ruppin/tiangen/CancerProject/05.scRNAseq_project/03.Results/seu_tumor_CD45_R_NR.rds")

tumor_CD45_cells_R_NR$major_cell_type <- tumor_CD45_cells_R_NR$DICE.fine
tumor_CD45_cells_R_NR$major_cell_type[tumor_CD45_cells_R_NR$major_cell_type%in%c("T cells, CD4+, Th1", "T cells, CD4+, Th17","T cells, CD4+, naive", "T cells, CD4+, naive, stimulated", "T cells, CD4+, memory TREG", "T cells, CD4+, TFH", "T cells, CD4+, Th2", "T cells, CD4+, Th1_17", "T cells, CD4+, naive TREG")] <- "CD4T"
tumor_CD45_cells_R_NR$major_cell_type[tumor_CD45_cells_R_NR$major_cell_type%in%c("T cells, CD8+, naive, stimulated", "T cells, CD8+, naive")] <- "CD8T"



```

# Plot UMAP of R and NR cells
```{r}
#### plot singleR annotation
DimPlot(tumor_CD45_cells_R_NR, group.by = "major_cell_type", label = TRUE, repel = TRUE, label.size = 2)+NoLegend()+
  # scale_color_manual(values = my_cols)+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_tumor_CD45_cells_R_NR_singler.png"), height = 120, width = 140, dpi = 800, units = "mm")

# BiocManager::install("dittoSeq")
library(dittoSeq)
dittoDimPlot(tumor_CD45_cells_R_NR, var = "major_cell_type", split.by = "predicted_label", size = 0.1)+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_tumor_CD45_cells_R_NR_predicted_label_split.png"), height = 120, width = 280, dpi = 800, units = "mm")
dittoBarPlot(tumor_CD45_cells_R_NR, var = "Volumetric_response", group.by = "DICE.fine")+ # predicted_label
  labs(x = "")+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_tumor_CD45_cells_R_NR_Volumetric_response_abundance.png"), height = 90, width = 200, dpi = 800, units = "mm")
```

# SingleR for tumor_CD45_cells
```{r}
ref <- celldex::DatabaseImmuneCellExpressionData() # celldex::BlueprintEncodeData()
singleR <- as.SingleCellExperiment(tumor_CD45_cells, assay= "RNA") 
singler.anno <- SingleR(test = singleR, ref = ref, labels = (ref$label.fine))
tumor_CD45_cells[["DICE.fine"]] <- singler.anno$labels # Database of Immune Cell Expression (DICE).
tumor_CD45_cells[["DICE.fine.pruned"]] <- singler.anno$pruned.labels
dittoBarPlot(tumor_CD45_cells, var = "Volumetric_response", group.by = "DICE.fine.pruned")+ # predicted_label
  labs(x = "")+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_tumor_CD45_cells_Volumetric_response_abundance.png"), height = 90, width = 200, dpi = 800, units = "mm")
saveRDS(object = tumor_CD45_cells, file = "/data/Lab_ruppin/tiangen/CancerProject/05.scRNAseq_project/03.Results/seu_tumor_CD45_singler_GSE200996.rds") # Here all genes are saves
```

# Analyze PENCIL identified phenotypic cells for blood samples
```{r}

predicted_cell_labels = read.csv("/data/Lab_ruppin/tiangen/CancerProject/05.scRNAseq_project/03.Results/PENCIL/GSE200996/py/Blood_Volumetric_response/predicted_labels.csv") # Blood_Volumetric_response   Volumetric_response
predicted_cell_labels_new <- subset(predicted_cell_labels, select = 'predicted_label')
meta <- blood_cells@meta.data
meta.new <- cbind(meta, predicted_cell_labels_new) # cell ID in PENCIL output does not match the original ones. But it is in the same order. So here we directly cbind two data frames to add the predicted label information for all cells.
blood_cells <- AddMetaData(blood_cells, meta.new)

```

# Plot predicted labels to confirm 
```{r}
DimPlot(blood_cells, group.by = "predicted_label")+
  # scale_color_manual(values = my.cols)+
  ggtitle("")
ggsave(filename = paste0(fig_dir, "UMAP_Blood_predicted_label.png"), height = 120, width = 190, dpi = 800, units = "mm")
```

# Calculate R/NR cell number and ratios in each patient
```{r}
# Calculate the cell number for each patient
grouped <- meta.new %>%
  group_by(meta, predicted_label)

# Calculate the number of cells for each patient and predicted_label
cell_counts <- grouped %>%
  summarise(cell_count = n())
#cell_counts <- cell_counts[cell_counts$predicted_label != "Rejected", ]

# Create a vector with all unique patient IDs
all_patients <- unique(cell_counts$meta)
# Create a new data frame with all patients having both 'yes' and 'no'
new_df <- data.frame(meta = rep(all_patients, each = 3),
                     predicted_label = rep(c("yes", "no", "Rejected"), length(all_patients)))
# Merge the new data frame with the original data frame based on patient ID and 'yes_no'
cell_counts_new <- merge(cell_counts, new_df, by = c("meta", "predicted_label"), all = TRUE)
# Replace missing values (NA) in the 'values' column with 0
cell_counts_new$cell_count[is.na(cell_counts_new$cell_count)] <- 0


# Calculate the total count for each patient
df_totals <- aggregate(cell_count ~ meta, cell_counts_new, sum)
# Calculate the count ratios for 'yes' and 'no'
df_ratios <- aggregate(cell_count ~ meta + predicted_label, cell_counts_new, sum)
df_ratios <- transform(df_ratios, ratio = cell_count / df_totals$cell_count * 100)

# Plot the stacked bar plot
ggplot(df_ratios, aes(x = meta, y = ratio, fill = predicted_label)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("yes" = "blue", "no" = "red")) +
  labs(x = "Patient", y = "Count Ratio", fill = "Response") +
  ggtitle("Count Ratios of 'R' and 'NR' Cells for Each Patient") +
  theme_minimal()
ggsave(filename = paste0(fig_dir, "Blood_R_NR_cellRatio_stackedBar.png"), height = 120, width = 190, dpi = 800, units = "mm")
```