---
title: "Pre-process the data for PENCIL input"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r}
# clear all objects including hidden objects
rm(list = ls(all.names = TRUE)) 
# free up memory and report the memory usage
gc()
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
require(ggplot2)
ggplot2::theme_set(new = theme_bw(base_size = 12)) # sets the theme and axis font size
```

# Connect to server
```{bash}
ssh changt7@biowulf.nih.gov
sinteractive --mem=64g --cpus-per-task=32 --time=24:00:00
sinteractive --mem=128g --cpus-per-task=32 --gres=gpu:k80:2 --time=24:00:00
# --gres=lscratch:1024
module load R/
R

```

# load required package
```{r, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("UCell")
library(UCell)

#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)

library(caret)
library(Seurat)
library(GEOquery)
library(dplyr)
library(ggplot2)
library(SingleR)
library(sctransform)
library(celldex)
library(dittoSeq)
library(SeuratWrappers)
library(harmony)
library(scuttle)
library(scran)
library(scFeatureFilter)
library(cowplot)
library(patchwork)
library(edgeR)
library(RColorBrewer)
library(stringr)

```


# set input and output directories & parameters
```{r}

# server
shared_data_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/02.Input/")
PENCIL_result_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/01.Scripts/results/")

dataset_train = 'GSE139324'
processed_dir_train <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_train, '/')
fig_dir_train = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_train, "/Figures/")

dataset_test1 = 'GSE180268'
processed_dir_test1 <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test1, '/')
fig_dir_test1 = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test1, "/Figures/")

dataset_test2 = 'GSE182227'
processed_dir_test2 <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test2, '/')
fig_dir_test2 = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test2, "/Figures/")

dataset_test3 = 'GSE200996'
processed_dir_test3 <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test3, '/')
fig_dir_test3 = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test3, "/Figures/")


```

# make PENCIL input files for training data (tumor CD8T cells)
```{r}

sampleType_name = 'Tumor'
phenotype_name = 'HPV'
cellType_of_interest = 'CD8T'

merged_seu_subset = readRDS(file = paste0(processed_dir_train,"seu_",sampleType_name,"_",cellType_of_interest,".rds"))

#### print cell number for each phenotype, used for class_weights setting
print(table(merged_seu_subset$HPV))

# filter genes to keep only protein coding genes
protein_coding_gene_file = paste0(shared_data_dir, 'Feldman_protein_coding_genes.csv')
protein_coding_genes=read.csv(protein_coding_gene_file, header=T)[,1]
genes <- intersect(protein_coding_genes, row.names(merged_seu_subset))
merged_seu_subset <- merged_seu_subset[genes, ]

merged_seu_subset <- FindVariableFeatures(object=merged_seu_subset, selection.method='vst', nfeatures=2000)
### default gene order by variance (high to low)
variable_genes <- VariableFeatures(merged_seu_subset)
merged_seu_subset <- ScaleData(merged_seu_subset)
merged_seu_subset <- RunPCA(object=merged_seu_subset, features=variable_genes)
merged_seu_subset <- FindNeighbors(object=merged_seu_subset, dims=1:10, k.param=20)
merged_seu_subset <- FindClusters(object=merged_seu_subset, resolution=1.6)
merged_seu_subset <- RunUMAP(object=merged_seu_subset, dims=1:10)


exp_data_scaled = merged_seu_subset@assays[["RNA"]]@scale.data[variable_genes,]
write.csv(exp_data_scaled, file=paste0(processed_dir_train,"seu_",sampleType_name,'_',cellType_of_interest,".csv"))
cell_label_df = data.frame(label = merged_seu_subset@meta.data[phenotype_name])
write.csv(cell_label_df, file=paste0(processed_dir_train,"cell_label_info.csv"))
emd = merged_seu_subset@reductions[["umap"]]@cell.embeddings
write.csv(emd, file=paste0(processed_dir_train, "seu_",sampleType_name,'_',cellType_of_interest,'_embedding_umap.csv'))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_train, "seu_",sampleType_name,"_", cellType_of_interest, "_PENCIL.rds")) 

```

## make PENCIL input files for test1 (Fill 0-value for missing genes in new test data)
```{r}

sampleType_name = 'Tumor'
phenotype_name = 'HPV'
cellType_of_interest = 'CD8T'

merged_seu_subset = readRDS(file = paste0(processed_dir_test1,"seu_",sampleType_name,"_",cellType_of_interest,".rds"))

num_fill_genes = length(variable_genes) - sum(variable_genes %in% row.names(merged_seu_subset))
print(num_fill_genes)
if (num_fill_genes > 0){
  num_cells = dim(merged_seu_subset)[2]
  fill_data <- rep(0, num_fill_genes * num_cells)
  dim(fill_data) <- c(num_fill_genes, num_cells)
  row.names(fill_data) <- setdiff(variable_genes, row.names(merged_seu_subset))
  colnames(fill_data) <- colnames(merged_seu_subset)
  counts <- as.matrix(merged_seu_subset@assays[["RNA"]]@counts)
  counts <- rbind(counts, fill_data)
  merged_seu_subset.filled <- CreateSeuratObject(counts=counts, meta.data=merged_seu_subset@meta.data)
  #merged_seu_subset.filled@reductions[["raw.umap"]] <- merged_seu_subset@reductions[["umap"]]
  merged_seu_subset = merged_seu_subset.filled
}
VariableFeatures(merged_seu_subset) <- variable_genes
merged_seu_subset <- NormalizeData(object=merged_seu_subset, normalization.method="LogNormalize")
merged_seu_subset <- ScaleData(object=merged_seu_subset)
merged_seu_subset <- RunPCA(object=merged_seu_subset, features=variable_genes)
merged_seu_subset <- FindNeighbors(object=merged_seu_subset, dims=1:10, k.param=20)
merged_seu_subset <- FindClusters(object=merged_seu_subset, resolution=1.6)
merged_seu_subset <- RunUMAP(object=merged_seu_subset, dims=1:10)


exp_data_scaled = merged_seu_subset@assays[["RNA"]]@scale.data[variable_genes,]
write.csv(exp_data_scaled, file=paste0(processed_dir_test1,"seu_",sampleType_name,'_',cellType_of_interest,".csv"))
cell_label_df = data.frame(label = merged_seu_subset@meta.data[phenotype_name])
write.csv(cell_label_df, file=paste0(processed_dir_test1,"cell_label_info.csv"))
emd = merged_seu_subset@reductions[["umap"]]@cell.embeddings
write.csv(emd, file=paste0(processed_dir_test1, "seu_",sampleType_name,'_',cellType_of_interest,'_embedding_umap.csv'))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test1, "seu_",sampleType_name,"_", cellType_of_interest, "_PENCIL.rds")) 

```

## make PENCIL input files for test2 (Fill 0-value for missing genes in new test data)
```{r}

sampleType_name = 'Tissue'
phenotype_name = 'HPV'
cellType_of_interest = 'CD8T'

merged_seu_subset = readRDS(file = paste0(processed_dir_test2,"seu_",sampleType_name,"_",cellType_of_interest,".rds"))

#merged_seu_subset$HPV[merged_seu_subset$HPV=='HPV+'] = 1
#merged_seu_subset$HPV[merged_seu_subset$HPV=='HPV-'] = 0
#saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test2, "seu_",sampleType_name,"_", cellType_of_interest, ".rds")) 


num_fill_genes = length(variable_genes) - sum(variable_genes %in% row.names(merged_seu_subset))
print(num_fill_genes)
if (num_fill_genes > 0){
  num_cells = dim(merged_seu_subset)[2]
  fill_data <- rep(0, num_fill_genes * num_cells)
  dim(fill_data) <- c(num_fill_genes, num_cells)
  row.names(fill_data) <- setdiff(variable_genes, row.names(merged_seu_subset))
  colnames(fill_data) <- colnames(merged_seu_subset)
  counts <- as.matrix(merged_seu_subset@assays[["RNA"]]@counts)
  counts <- rbind(counts, fill_data)
  merged_seu_subset.filled <- CreateSeuratObject(counts=counts, meta.data=merged_seu_subset@meta.data)
  #merged_seu_subset.filled@reductions[["raw.umap"]] <- merged_seu_subset@reductions[["umap"]]
  merged_seu_subset = merged_seu_subset.filled
}
VariableFeatures(merged_seu_subset) <- variable_genes
merged_seu_subset <- NormalizeData(object=merged_seu_subset, normalization.method="LogNormalize")
merged_seu_subset <- ScaleData(object=merged_seu_subset)
merged_seu_subset <- RunPCA(object=merged_seu_subset, features=variable_genes)
merged_seu_subset <- FindNeighbors(object=merged_seu_subset, dims=1:10, k.param=20)
merged_seu_subset <- FindClusters(object=merged_seu_subset, resolution=1.6)
merged_seu_subset <- RunUMAP(object=merged_seu_subset, dims=1:10)


exp_data_scaled = merged_seu_subset@assays[["RNA"]]@scale.data[variable_genes,]
write.csv(exp_data_scaled, file=paste0(processed_dir_test2,"seu_",sampleType_name,'_',cellType_of_interest,".csv"))
cell_label_df = data.frame(label = merged_seu_subset@meta.data[phenotype_name])
write.csv(cell_label_df, file=paste0(processed_dir_test2,"cell_label_info.csv"))
emd = merged_seu_subset@reductions[["umap"]]@cell.embeddings
write.csv(emd, file=paste0(processed_dir_test2, "seu_",sampleType_name,'_',cellType_of_interest,'_embedding_umap.csv'))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test2, "seu_",sampleType_name,"_", cellType_of_interest, "_PENCIL.rds")) 

```


## make PENCIL input files for test3 (Fill 0-value for missing genes in new test data)
```{r}

sampleType_name = 'Tissue'
phenotype_name = 'HPV'
cellType_of_interest = 'CD8T'

merged_seu_subset = readRDS(file = paste0(processed_dir_test3,"seu_",sampleType_name,"_",cellType_of_interest,".rds"))

#merged_seu_subset$HPV = 0
#saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test3, "seu_",sampleType_name,"_", cellType_of_interest, ".rds")) 

num_fill_genes = length(variable_genes) - sum(variable_genes %in% row.names(merged_seu_subset))
print(num_fill_genes)
if (num_fill_genes > 0){
  num_cells = dim(merged_seu_subset)[2]
  fill_data <- rep(0, num_fill_genes * num_cells)
  dim(fill_data) <- c(num_fill_genes, num_cells)
  row.names(fill_data) <- setdiff(variable_genes, row.names(merged_seu_subset))
  colnames(fill_data) <- colnames(merged_seu_subset)
  counts <- as.matrix(merged_seu_subset@assays[["RNA"]]@counts)
  counts <- rbind(counts, fill_data)
  merged_seu_subset.filled <- CreateSeuratObject(counts=counts, meta.data=merged_seu_subset@meta.data)
  #merged_seu_subset.filled@reductions[["raw.umap"]] <- merged_seu_subset@reductions[["umap"]]
  merged_seu_subset = merged_seu_subset.filled
}
VariableFeatures(merged_seu_subset) <- variable_genes
merged_seu_subset <- NormalizeData(object=merged_seu_subset, normalization.method="LogNormalize")
merged_seu_subset <- ScaleData(object=merged_seu_subset)
merged_seu_subset <- RunPCA(object=merged_seu_subset, features=variable_genes)
merged_seu_subset <- FindNeighbors(object=merged_seu_subset, dims=1:10, k.param=20)
merged_seu_subset <- FindClusters(object=merged_seu_subset, resolution=1.6)
merged_seu_subset <- RunUMAP(object=merged_seu_subset, dims=1:10)


exp_data_scaled = merged_seu_subset@assays[["RNA"]]@scale.data[variable_genes,]
write.csv(exp_data_scaled, file=paste0(processed_dir_test3,"seu_",sampleType_name,'_',cellType_of_interest,".csv"))
cell_label_df = data.frame(label = merged_seu_subset@meta.data[phenotype_name])
write.csv(cell_label_df, file=paste0(processed_dir_test3,"cell_label_info.csv"))
emd = merged_seu_subset@reductions[["umap"]]@cell.embeddings
write.csv(emd, file=paste0(processed_dir_test3, "seu_",sampleType_name,'_',cellType_of_interest,'_embedding_umap.csv'))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test3, "seu_",sampleType_name,"_", cellType_of_interest, "_PENCIL.rds")) 

```

# Run PENCIL
```{bash}

## one must use gpu to accelerate running PENCIL, otherwise would be extremely slow
sinteractive --mem=128g --cpus-per-task=32 --gres=gpu:k80:2 --time=24:00:00
## PENCIL has been installed in python/3.8
module load python/3.8

## Run PENCIL with .csv files as input, which seems work well
python 02.Run_PENCIL_csv.py /data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/GSE139324/seu_Tumor_CD8T.csv /data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/GSE139324/cell_label_info.csv /data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/GSE139324/seu_Tumor_CD8T_embedding_umap.csv GSE139324_Tumor_CD8T HPVinfection multi-classification 1,2 /data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/GSE180268/seu_Tumor_CD8T.csv,/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/GSE182227/seu_Tissue_CD8T.csv,/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/GSE200996/seu_Tissue_CD8T.csv

```


# Add PENCIL identified phenotypic cell labels to all training and test datasets
```{r}

#### for training data
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_train,"seu_Tumor_CD8T.rds"))

predicted_cell_labels = read.csv(paste0(PENCIL_result_dir, 'GSE139324_Tumor_CD8T/py/HPVinfection/predicted_labels.csv'))
predicted_cell_labels_new <- subset(predicted_cell_labels, select = 'predicted_label')
colnames(predicted_cell_labels_new) = c("PENCIL_pred_label")

# change label names to yes, no, rej
label_names = unique(predicted_cell_labels_new$PENCIL_pred_label)
print(label_names)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("1", "yes", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("0", "no", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("Rejected", "rej", predicted_cell_labels_new$PENCIL_pred_label)

meta <- merged_seu_subset@meta.data
meta.new <- cbind(meta, predicted_cell_labels_new) # cell ID in PENCIL output does not match the original ones. But it is in the same order.
merged_seu_subset <- AddMetaData(merged_seu_subset, meta.new)
merged_seu_subset$PENCIL_pred_label <- factor(merged_seu_subset$PENCIL_pred_label, levels = c("yes", "no", "rej"))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_train,"seu_Tumor_CD8T_HPV_PENCIL.rds")) 


#### for test1 data
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_test1,"seu_Tumor_CD8T.rds"))

predicted_cell_labels = read.csv(paste0(PENCIL_result_dir, 'GSE139324_Tumor_CD8T/py/HPVinfection/predicted_labels_test1.csv'))
predicted_cell_labels_new <- subset(predicted_cell_labels, select = 'predicted_label')
colnames(predicted_cell_labels_new) = c("PENCIL_pred_label")

# change label names to yes, no, rej
label_names = unique(predicted_cell_labels_new$PENCIL_pred_label)
print(label_names)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("1", "yes", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("0", "no", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("Rejected", "rej", predicted_cell_labels_new$PENCIL_pred_label)

meta <- merged_seu_subset@meta.data
meta.new <- cbind(meta, predicted_cell_labels_new) # cell ID in PENCIL output does not match the original ones. But it is in the same order.
merged_seu_subset <- AddMetaData(merged_seu_subset, meta.new)
merged_seu_subset$PENCIL_pred_label <- factor(merged_seu_subset$PENCIL_pred_label, levels = c("yes", "no", "rej"))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test1,"seu_Tumor_CD8T_HPV_PENCIL.rds")) 


#### for test2 data
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_test2,"seu_Tissue_CD8T.rds"))

predicted_cell_labels = read.csv(paste0(PENCIL_result_dir, 'GSE139324_Tumor_CD8T/py/HPVinfection/predicted_labels_test2.csv'))
predicted_cell_labels_new <- subset(predicted_cell_labels, select = 'predicted_label')
colnames(predicted_cell_labels_new) = c("PENCIL_pred_label")

# change label names to yes, no, rej
label_names = unique(predicted_cell_labels_new$PENCIL_pred_label)
print(label_names)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("1", "yes", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("0", "no", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("Rejected", "rej", predicted_cell_labels_new$PENCIL_pred_label)

meta <- merged_seu_subset@meta.data
meta.new <- cbind(meta, predicted_cell_labels_new) # cell ID in PENCIL output does not match the original ones. But it is in the same order.
merged_seu_subset <- AddMetaData(merged_seu_subset, meta.new)
merged_seu_subset$PENCIL_pred_label <- factor(merged_seu_subset$PENCIL_pred_label, levels = c("yes", "no", "rej"))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test2,"seu_Tissue_CD8T.rds")) 


#### for test3 data
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_test3,"seu_Tissue_CD8T.rds"))

predicted_cell_labels = read.csv(paste0(PENCIL_result_dir, 'GSE139324_Tumor_CD8T/py/HPVinfection/predicted_labels_test3.csv'))
predicted_cell_labels_new <- subset(predicted_cell_labels, select = 'predicted_label')
colnames(predicted_cell_labels_new) = c("PENCIL_pred_label")

# change label names to yes, no, rej
label_names = unique(predicted_cell_labels_new$PENCIL_pred_label)
print(label_names)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("1", "yes", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("0", "no", predicted_cell_labels_new$PENCIL_pred_label)
predicted_cell_labels_new$PENCIL_pred_label <- gsub("Rejected", "rej", predicted_cell_labels_new$PENCIL_pred_label)

meta <- merged_seu_subset@meta.data
meta.new <- cbind(meta, predicted_cell_labels_new) # cell ID in PENCIL output does not match the original ones. But it is in the same order.
merged_seu_subset <- AddMetaData(merged_seu_subset, meta.new)
merged_seu_subset$PENCIL_pred_label <- factor(merged_seu_subset$PENCIL_pred_label, levels = c("yes", "no", "rej"))

saveRDS(object = merged_seu_subset, file = paste0(processed_dir_test3,"seu_Tissue_CD8T.rds")) 

```


# Calculate R/NR cell number and ratios in each patient for all training and test datasets
```{r}

min_cellNum_per_patient = 50

#### for training data
phenotype_name = 'HPV'
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_train,"seu_Tumor_CD8T_HPV_PENCIL.rds"))
merged_seu_subset$meta = merged_seu_subset$patient
patient_CD8Tcell_num = sort(table(merged_seu_subset$meta))
keep_patient = names(patient_CD8Tcell_num)[patient_CD8Tcell_num > min_cellNum_per_patient]
merged_seu_subset = subset(merged_seu_subset, subset = meta %in% keep_patient)
sampleNUM = length(unique(merged_seu_subset$meta))
print(paste('Total samples: ', sampleNUM))

# Calculate the cell number for each patient
grouped <- merged_seu_subset@meta.data %>% group_by(meta, PENCIL_pred_label)
# Calculate the number of cells for each patient and predicted_label
cell_counts <- grouped %>% summarise(cell_count = n())
cell_counts <- cell_counts[cell_counts$PENCIL_pred_label != "rej", ]
# Create a vector with all unique patient IDs
all_patients <- unique(cell_counts$meta)
print(unique(merged_seu_subset$PENCIL_pred_label))

unique_label = c("yes","no")
# Create a new data frame with all patients having both 'yes' and 'no'
new_df <- data.frame(meta = rep(all_patients, each = 2),
                     PENCIL_pred_label = rep(unique_label, length(all_patients)))
# Merge the new data frame with the original data frame based on patient ID and 'yes_no'
cell_counts_new <- merge(cell_counts, new_df, by = c("meta", "PENCIL_pred_label"), all = TRUE)
# Replace missing values (NA) in the 'values' column with 0
cell_counts_new$cell_count[is.na(cell_counts_new$cell_count)] <- 0
# Calculate the total count for each patient
df_totals <- aggregate(cell_count ~ meta, cell_counts_new, sum)
# Calculate the count ratios for 'yes' and 'no'
df_ratios <- aggregate(cell_count ~ meta + PENCIL_pred_label, cell_counts_new, sum)
df_ratios <- transform(df_ratios, ratio = cell_count / df_totals$cell_count * 100)

phenotype_df = merged_seu_subset@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
# keep only non-duplicated patientID --> dim(phenotype_df.new)=c(48,3)
df_ratios <- merge(df_ratios, phenotype_df, by.x = "meta", by.y = "meta")
df_ratios <- df_ratios[order(df_ratios[,phenotype_name]), ]
df_ratios$meta <- factor(df_ratios$meta, levels = unique(df_ratios$meta))
print(paste('HPV+ sample number: ', sum(df_ratios$HPV==1)/2))

df_ratios_yes = df_ratios[df_ratios$PENCIL_pred_label=='yes',]
ratios_R = df_ratios_yes$ratio[df_ratios_yes$HPV==1]
ratios_NR = df_ratios_yes$ratio[df_ratios_yes$HPV==0]
result <- wilcox.test(ratios_R, ratios_NR, alternative = "greater")
print(paste("Hypothesis R > NR: ",result$p.value))

roc_curve <- roc(df_ratios_yes$HPV, df_ratios_yes$ratio)
auc_value <- auc(roc_curve)
acc_value = mean(df_ratios_yes$HPV == as.numeric(df_ratios_yes$ratio>50))
print(paste('AUC = ', auc_value, '. ACC = ', acc_value))

# Plot the stacked bar plot
ggplot(df_ratios, aes(x = meta, y = ratio, fill = PENCIL_pred_label)) +
  geom_bar(stat = "identity") +
  labs(x = NULL, y = "Percentage (%)", fill = "PENCIL-pred") +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.x = element_blank(),  # Removes x-axis labels
    axis.ticks.x = element_blank(), # Removes x-axis tick marks
    legend.position = "none"       # Removes legend
  ) # +
  #scale_fill_manual(values = c("yes" = "#A6CEE3FF", "no" = "#5AA3DAFF"))
ggsave(filename = paste0(PENCIL_result_dir, "GSE139324_Tumor_CD8T/py/HPVinfection/stackedBar_cellRatio_train.pdf"), height = 40, width = 15+25*sampleNUM/15, units = "mm")




#### for test1 data
phenotype_name = 'HPV'
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_test1,"seu_Tumor_CD8T.rds"))
patient_CD8Tcell_num = sort(table(merged_seu_subset$meta))
keep_patient = names(patient_CD8Tcell_num)[patient_CD8Tcell_num > min_cellNum_per_patient]
merged_seu_subset = subset(merged_seu_subset, subset = meta %in% keep_patient)
sampleNUM = length(unique(merged_seu_subset$meta))
print(paste('Total samples: ', sampleNUM))

# Calculate the cell number for each patient
grouped <- merged_seu_subset@meta.data %>% group_by(meta, PENCIL_pred_label)
# Calculate the number of cells for each patient and predicted_label
cell_counts <- grouped %>% summarise(cell_count = n())
cell_counts <- cell_counts[cell_counts$PENCIL_pred_label != "rej", ]
# Create a vector with all unique patient IDs
all_patients <- unique(cell_counts$meta)
print(unique(merged_seu_subset$PENCIL_pred_label))

unique_label = c("yes","no")
# Create a new data frame with all patients having both 'yes' and 'no'
new_df <- data.frame(meta = rep(all_patients, each = 2),
                     PENCIL_pred_label = rep(unique_label, length(all_patients)))
# Merge the new data frame with the original data frame based on patient ID and 'yes_no'
cell_counts_new <- merge(cell_counts, new_df, by = c("meta", "PENCIL_pred_label"), all = TRUE)
# Replace missing values (NA) in the 'values' column with 0
cell_counts_new$cell_count[is.na(cell_counts_new$cell_count)] <- 0
# Calculate the total count for each patient
df_totals <- aggregate(cell_count ~ meta, cell_counts_new, sum)
# Calculate the count ratios for 'yes' and 'no'
df_ratios <- aggregate(cell_count ~ meta + PENCIL_pred_label, cell_counts_new, sum)
df_ratios <- transform(df_ratios, ratio = cell_count / df_totals$cell_count * 100)

phenotype_df = merged_seu_subset@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
# keep only non-duplicated patientID --> dim(phenotype_df.new)=c(48,3)
df_ratios <- merge(df_ratios, phenotype_df, by.x = "meta", by.y = "meta")
df_ratios <- df_ratios[order(df_ratios[,phenotype_name]), ]
df_ratios$meta <- factor(df_ratios$meta, levels = unique(df_ratios$meta))
print(paste('HPV+ sample number: ', sum(df_ratios$HPV==1)/2))

df_ratios_yes = df_ratios[df_ratios$PENCIL_pred_label=='yes',]
ratios_R = df_ratios_yes$ratio[df_ratios_yes$HPV==1]
ratios_NR = df_ratios_yes$ratio[df_ratios_yes$HPV==0]
result <- wilcox.test(ratios_R, ratios_NR, alternative = "greater")
print(paste("Hypothesis R > NR: ",result$p.value))

roc_curve <- roc(df_ratios_yes$HPV, df_ratios_yes$ratio)
auc_value <- auc(roc_curve)
acc_value = mean(df_ratios_yes$HPV == as.numeric(df_ratios_yes$ratio>50))
print(paste('AUC = ', auc_value, '. ACC = ', acc_value))

# Plot the stacked bar plot
ggplot(df_ratios, aes(x = meta, y = ratio, fill = PENCIL_pred_label)) +
  geom_bar(stat = "identity") +
  labs(x = NULL, y = "Percentage (%)", fill = "PENCIL-pred") +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.x = element_blank(),  # Removes x-axis labels
    axis.ticks.x = element_blank(), # Removes x-axis tick marks
    legend.position = "none"       # Removes legend
  ) # +
  #scale_fill_manual(values = c("yes" = "#A6CEE3FF", "no" = "#5AA3DAFF"))
ggsave(filename = paste0(PENCIL_result_dir, "GSE139324_Tumor_CD8T/py/HPVinfection/stackedBar_cellRatio_test1.pdf"), height = 40, width = 15+25*sampleNUM/15, units = "mm")





#### for test2 data
phenotype_name = 'HPV'
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_test2,"seu_Tissue_CD8T.rds"))
patient_CD8Tcell_num = sort(table(merged_seu_subset$meta))
keep_patient = names(patient_CD8Tcell_num)[patient_CD8Tcell_num > min_cellNum_per_patient]
merged_seu_subset = subset(merged_seu_subset, subset = meta %in% keep_patient)
sampleNUM = length(unique(merged_seu_subset$meta))
print(paste('Total samples: ', sampleNUM))

# Calculate the cell number for each patient
grouped <- merged_seu_subset@meta.data %>% group_by(meta, PENCIL_pred_label)
# Calculate the number of cells for each patient and predicted_label
cell_counts <- grouped %>% summarise(cell_count = n())
cell_counts <- cell_counts[cell_counts$PENCIL_pred_label != "rej", ]
# Create a vector with all unique patient IDs
all_patients <- unique(cell_counts$meta)
print(unique(merged_seu_subset$PENCIL_pred_label))

unique_label = c("yes","no")
# Create a new data frame with all patients having both 'yes' and 'no'
new_df <- data.frame(meta = rep(all_patients, each = 2),
                     PENCIL_pred_label = rep(unique_label, length(all_patients)))
# Merge the new data frame with the original data frame based on patient ID and 'yes_no'
cell_counts_new <- merge(cell_counts, new_df, by = c("meta", "PENCIL_pred_label"), all = TRUE)
# Replace missing values (NA) in the 'values' column with 0
cell_counts_new$cell_count[is.na(cell_counts_new$cell_count)] <- 0
# Calculate the total count for each patient
df_totals <- aggregate(cell_count ~ meta, cell_counts_new, sum)
# Calculate the count ratios for 'yes' and 'no'
df_ratios <- aggregate(cell_count ~ meta + PENCIL_pred_label, cell_counts_new, sum)
df_ratios <- transform(df_ratios, ratio = cell_count / df_totals$cell_count * 100)

phenotype_df = merged_seu_subset@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
# keep only non-duplicated patientID --> dim(phenotype_df.new)=c(48,3)
df_ratios <- merge(df_ratios, phenotype_df, by.x = "meta", by.y = "meta")
df_ratios <- df_ratios[order(df_ratios[,phenotype_name]), ]
df_ratios$meta <- factor(df_ratios$meta, levels = unique(df_ratios$meta))
print(paste('HPV+ sample number: ', sum(df_ratios$HPV==1)/2))

# Plot the stacked bar plot
ggplot(df_ratios, aes(x = meta, y = ratio, fill = PENCIL_pred_label)) +
  geom_bar(stat = "identity") +
  labs(x = NULL, y = "Percentage (%)", fill = "PENCIL-pred") +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.x = element_blank(),  # Removes x-axis labels
    axis.ticks.x = element_blank(), # Removes x-axis tick marks
    legend.position = "none"       # Removes legend
  ) # +
  #scale_fill_manual(values = c("yes" = "#A6CEE3FF", "no" = "#5AA3DAFF"))
ggsave(filename = paste0(PENCIL_result_dir, "GSE139324_Tumor_CD8T/py/HPVinfection/stackedBar_cellRatio_test2.pdf"), height = 40, width = 15+25*sampleNUM/15, units = "mm")




#### for test3 data
phenotype_name = 'HPV'
# load seurat object
merged_seu_subset = readRDS(file = paste0(processed_dir_test3,"seu_Tissue_CD8T.rds"))
patient_CD8Tcell_num = sort(table(merged_seu_subset$meta))
keep_patient = names(patient_CD8Tcell_num)[patient_CD8Tcell_num > min_cellNum_per_patient]
merged_seu_subset = subset(merged_seu_subset, subset = meta %in% keep_patient)
sampleNUM = length(unique(merged_seu_subset$meta))
print(paste('Total samples: ', sampleNUM))

# Calculate the cell number for each patient
grouped <- merged_seu_subset@meta.data %>% group_by(meta, PENCIL_pred_label)
# Calculate the number of cells for each patient and predicted_label
cell_counts <- grouped %>% summarise(cell_count = n())
cell_counts <- cell_counts[cell_counts$PENCIL_pred_label != "rej", ]
# Create a vector with all unique patient IDs
all_patients <- unique(cell_counts$meta)
print(unique(merged_seu_subset$PENCIL_pred_label))

unique_label = c("yes","no")
# Create a new data frame with all patients having both 'yes' and 'no'
new_df <- data.frame(meta = rep(all_patients, each = 2),
                     PENCIL_pred_label = rep(unique_label, length(all_patients)))
# Merge the new data frame with the original data frame based on patient ID and 'yes_no'
cell_counts_new <- merge(cell_counts, new_df, by = c("meta", "PENCIL_pred_label"), all = TRUE)
# Replace missing values (NA) in the 'values' column with 0
cell_counts_new$cell_count[is.na(cell_counts_new$cell_count)] <- 0
# Calculate the total count for each patient
df_totals <- aggregate(cell_count ~ meta, cell_counts_new, sum)
# Calculate the count ratios for 'yes' and 'no'
df_ratios <- aggregate(cell_count ~ meta + PENCIL_pred_label, cell_counts_new, sum)
df_ratios <- transform(df_ratios, ratio = cell_count / df_totals$cell_count * 100)

phenotype_df = merged_seu_subset@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
# keep only non-duplicated patientID --> dim(phenotype_df.new)=c(48,3)
df_ratios <- merge(df_ratios, phenotype_df, by.x = "meta", by.y = "meta")
df_ratios <- df_ratios[order(df_ratios[,phenotype_name]), ]
df_ratios$meta <- factor(df_ratios$meta, levels = unique(df_ratios$meta))
print(paste('HPV+ sample number: ', sum(df_ratios$HPV==1)/2))

# Plot the stacked bar plot
ggplot(df_ratios, aes(x = meta, y = ratio, fill = PENCIL_pred_label)) +
  geom_bar(stat = "identity") +
  labs(x = NULL, y = "Percentage (%)", fill = "PENCIL-pred") +
  theme(
    panel.background = element_rect(fill = "white"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    #axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.x = element_blank(),  # Removes x-axis labels
    axis.ticks.x = element_blank(), # Removes x-axis tick marks
    legend.position = "none"       # Removes legend
  ) # +
  #scale_fill_manual(values = c("yes" = "#A6CEE3FF", "no" = "#5AA3DAFF"))
ggsave(filename = paste0(PENCIL_result_dir, "GSE139324_Tumor_CD8T/py/HPVinfection/stackedBar_cellRatio_test3.pdf"), height = 40, width = 15+25*sampleNUM/15, units = "mm")

```


# Use marker gene signature score to predict ICB response in 0.1) the same training data; 0.2) the same training data using blood-derived score; 1) bulk RNAseq; 2) other cancer types.
```{r}

merged_seu_subset$meta <- factor(merged_seu_subset$meta, levels = unique(df_ratios$meta0))

feats <- c("NR_cell_marker", "Rej_cell_marker", "R_cell_marker", "R_minus_NR_cell_marker")
p <- VlnPlot(merged_seu_subset, group.by = "meta", features = feats, pt.size = 0, ncol = 1) +
  NoLegend()
ggsave(plot = p,path = fig_dir,  filename = paste0("Vln_",sampleType_name,'_',cellType_of_interest,"_TopN.marker_patient.pdf"),  width = 14, height = 8)


saveRDS(object = merged_seu_subset, file = paste0(processed_dir, "seu_",sampleType_name,"_", cellType_of_interest, ".rds")) 

```

