---
title: "Pre-process the data for PENCIL input"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r}
# clear all objects including hidden objects
rm(list = ls(all.names = TRUE)) 
# free up memory and report the memory usage
gc()
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      comment = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
require(ggplot2)
ggplot2::theme_set(new = theme_bw(base_size = 12)) # sets the theme and axis font size
```

# Connect to server
```{bash}
ssh changt7@biowulf.nih.gov
sinteractive --mem=64g --cpus-per-task=32 --time=24:00:00
sinteractive --mem=128g --cpus-per-task=32 --gres=gpu:k80:2 --time=24:00:00
# --gres=lscratch:1024
module load R/
R

```

# load required package
```{r, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("UCell")
library(UCell)

#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)

library(caret)
library(Seurat)
library(GEOquery)
library(dplyr)
library(ggplot2)
library(SingleR)
library(sctransform)
library(celldex)
library(dittoSeq)
library(SeuratWrappers)
library(harmony)
library(scuttle)
library(scran)
library(scFeatureFilter)
library(cowplot)
library(patchwork)
library(edgeR)
library(RColorBrewer)
library(stringr)
#BiocManager::install("GSVA")
library(GSVA)
library(pROC)

```

# set input and output directories & parameters
```{r}

phenotype_name = 'ICBresponse'

# server
shared_data_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/02.Input/")
PENCIL_result_dir = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/01.Scripts/results/")
PENCIL_result_dir_deep = paste0(PENCIL_result_dir, 'GSE120575_Tissue_CD8T/py/',phenotype_name,'/')

dataset_train = 'GSE120575'
processed_dir_train <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_train, '/')
fig_dir_train = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_train, "/Figures/")

dataset_test1 = 'GSE123813'
processed_dir_test1 <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test1, '/')
fig_dir_test1 = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test1, "/Figures/")

dataset_test2 = 'GSE166181'
processed_dir_test2 <- paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test2, '/')
fig_dir_test2 = paste0("/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/", dataset_test2, "/Figures/")


```


# Identify DEGs for PENCIL N/NR cells on training data
```{r}

logFCfilter = 0.25 # 0.5
adjPvalFilter = 0.05

# loading seurat object of training data
merged_seu = readRDS(file = paste0(processed_dir_train,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))

##### find all markers for PENCIL predicted labels
Idents(object = merged_seu) <- merged_seu@meta.data$'PENCIL_pred_label_ICBresponse'
Cells.markers = FindAllMarkers(merged_seu, only.pos=T, min.pct=0.25, logfc.threshold = logFCfilter)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'

sig.markers = Cells.markers[(abs(as.numeric(as.vector(Cells.markers$avg_log2FC))) > logFCfilter & 
                               as.numeric(as.vector(Cells.markers$p_val_adj)) < adjPvalFilter), ]

write.csv(sig.markers, paste0(processed_dir, "sig.markers_PENCILpredLabel_",sampleType_name, '_', cellType_of_interest,".csv"))

##### Keep only the top N genes for each PENCIL predicted label, ranked by logFC
topN = 5
TopN.markers <- sig.markers %>%
    group_by(cluster) %>%
    top_n(topN, avg_log2FC)

R_markerGenes = TopN.markers$gene[TopN.markers$cluster=='yes']
NR_markerGenes = TopN.markers$gene[TopN.markers$cluster=='no']

### marker genes from PENCIL paper (much worse due to wrong processing of data)
# R_markerGenes = c("IL7R", "GPR183", "CCR7", "NR4A2", "LMNA", "NR4A3", "CD55", "MGAT4A", "PER1", "TSPYL2", "FOXP1", "STAT4")
# NR_markerGenes = c("CD38", "CCL3", "VCAM1", "HAVCR2", "CCR5", "GOLIM4", "PRDX3", "LGALS9", "ENTPD1", "TRAFD1", "PDCD1", "BATF", "LAG3", "CTLA4", "PTPN6", "CXCR6")
# R_markerGenes = R_markerGenes[1:5]
# NR_markerGenes = NR_markerGenes[1:5]
```

# plot dotplot for the topN marker genes
```{r}

merged_seu_sub = subset(merged_seu, subset = PENCIL_pred_label_ICBresponse != "rej")
merged_seu_sub$PENCIL_pred_label_ICBresponse = as.character(merged_seu_sub$PENCIL_pred_label_ICBresponse)
merged_seu_sub$PENCIL_pred_label_ICBresponse[merged_seu_sub$PENCIL_pred_label_ICBresponse=='no'] = 'Non-responder'
merged_seu_sub$PENCIL_pred_label_ICBresponse[merged_seu_sub$PENCIL_pred_label_ICBresponse=='yes'] = 'Responder'

p <- DotPlot(merged_seu_sub, 
             features = c(R_markerGenes,NR_markerGenes), 
             group.by = "PENCIL_pred_label_ICBresponse",
             cols = c("lightgrey", "red")) + 
  labs(x="", y="")+
  #coord_flip()+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 15),
        axis.text.y = element_text(color = "black", size = 15),
        legend.position = c(1.25, 0.3),
        legend.key.height = unit(5, "mm"),
        plot.margin = margin(t = 1, r = 40, b = 1, l = 1, unit = "mm"))
ggsave(filename = paste0(PENCIL_result_dir_deep, "Dotplot_TopNmarkers_train.pdf"), plot = p, height = 30*2.5, width = 40*4, units = "mm")


```


# Use marker gene GSVA score to predict ICB response in 1) the training set (using CD8T pseudo-bulk) and 2) test sets (using CD8T pseudo-bulk)
```{r}

min_cellNum_per_sample = 50

gene_set_list <- list(
  GeneSet1 = R_markerGenes,
  GeneSet2 = NR_markerGenes
)

result_list = list() # store GSVA score and ICB response info of different datasets

#### 1) the training set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_train,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# save to result
response <- ifelse(gsva_scores$ICBresponse > 0, 1, 0) 
predictor = as.numeric(gsva_scores$GSVA_diff)
result_list[['train']] = data.frame(ICBresponse = response, GSVAscore = predictor)



#### 2) the test1 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test1,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))
#merged_seu$meta = merged_seu$patient
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# save to result
response <- ifelse(gsva_scores$ICBresponse > 0, 1, 0)
predictor = as.numeric(gsva_scores$GSVA_diff)
result_list[['test1']] = data.frame(ICBresponse = response, GSVAscore = predictor)



#### 3) the test2 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test2,"seu_PBMC_CD8T_",phenotype_name,"_PENCIL.rds"))
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# save to result
response <- ifelse(gsva_scores$ICBresponse > 0, 1, 0)
predictor = as.numeric(gsva_scores$GSVA_diff)
result_list[['test2']] = data.frame(ICBresponse = response, GSVAscore = predictor)

ROC_list <- lapply(result_list, function(df) {
  roc_curve <- roc(df$ICBresponse, df$GSVAscore, grea) # , levels = c(0, 1), direction = "<"
  auc_value <- auc(roc_curve)
  acc_value = mean(df$ICBresponse == (df$GSVAscore>0))
  return(list(roc_curve = roc_curve, auc_value = auc_value, acc_value = acc_value))
})



#### Plot the ROC curves and AUC values
pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_ROC.pdf"),height = 2.5, width = 2.5) 
ggplot() +
  #theme_minimal() +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black"))+
  #theme(legend.position = "bottom") +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "1 - Specificity", y = "Sensitivity") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  geom_line(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "black", linetype = "solid", linewidth = 1) +
  geom_line(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#FD3A4A", linetype = "solid", linewidth = 1) +
  geom_line(data = data.frame(specificity=ROC_list[[3]]$roc_curve$sensitivities, sensitivity=ROC_list[[3]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#1F8A70", linetype = "solid", linewidth = 1) +
  geom_text(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 0.6, y = 0.3, label = paste("AUC (train) =", round(ROC_list[[1]]$auc_value, 2))), color = "black",fontface = "plain") +
  geom_text(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 0.6, y = 0.2, label = paste("AUC (test1) =", round(ROC_list[[2]]$auc_value, 2))), color = "#FD3A4A",fontface = "plain") +
  geom_text(data = data.frame(specificity=ROC_list[[3]]$roc_curve$sensitivities, sensitivity=ROC_list[[3]]$roc_curve$specificities), aes(x = 0.6, y = 0.1, label = paste("AUC (test2) =", round(ROC_list[[3]]$auc_value, 2))), color = "#1F8A70",fontface = "plain")#+
  #scale_color_manual(values = c("blue", "red", "green"), labels = c("Train", "Test1", "Test2"))
dev.off()



### boxplot of GSVA scores in R and NR samples
combined_df <- do.call(rbind, result_list)
pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_boxplot.pdf"),height = 2.5*0.7, width = 2.5*1) 
ggplot() +
  geom_boxplot(data = result_list[[1]], aes(x = factor(ICBresponse, level=c(1,0)), y = GSVAscore, fill = factor(ICBresponse, level=c(1,0)))) +
  geom_boxplot(data = result_list[[2]], aes(x = factor(ICBresponse+2, level=c(3,2)), y = GSVAscore, fill = factor(ICBresponse, level=c(1,0)))) +
  geom_boxplot(data = result_list[[3]], aes(x = factor(ICBresponse+4, level=c(5,4)), y = GSVAscore, fill = factor(ICBresponse, level=c(1,0)))) +
  labs(x = "", y = "Score") +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black"),
        legend.position = "none")
dev.off()

print(wilcox.test(result_list[[1]]$GSVAscore[result_list[[1]]$ICBresponse==1], result_list[[1]]$GSVAscore[result_list[[1]]$ICBresponse==0], alternative = "greater"))
print(wilcox.test(result_list[[2]]$GSVAscore[result_list[[2]]$ICBresponse==1], result_list[[2]]$GSVAscore[result_list[[2]]$ICBresponse==0], alternative = "greater"))
print(wilcox.test(result_list[[3]]$GSVAscore[result_list[[3]]$ICBresponse==1], result_list[[3]]$GSVAscore[result_list[[3]]$ICBresponse==0], alternative = "greater"))

```







# Use marker gene GSVA score to predict ICB response in other cancer types (using CD8T pseudo-bulk). Conlcusion: not work
```{r}

#### HNSCC GSE200996
phenotype_name = 'RECIST_response' # Any_response  RECIST_response
#### 4) the test4 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test4,"seu_Tissue_CD8T.rds"))
#merged_seu$meta = merged_seu$patient
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta
if (phenotype_name == "RECIST_response"){
  phenotype_df = subset(phenotype_df, subset = RECIST_response!="not measurable")
}

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# Calculate AUC using pROC package
if (phenotype_name == "RECIST_response"){
  response <- ifelse(gsva_scores[[phenotype_name]] == 'response', 1, 0) # RECIST
}else{
  response <- ifelse(gsva_scores[[phenotype_name]] == 'yes', 1, 0) # Anyreponse 
}
predictor = as.numeric(gsva_scores$GSVA_diff)
roc_obj <- roc(response, predictor, direction="<")
# Print the AUC value
auc_value <- auc(roc_obj)
print(paste("ICB prediction AUC for test4 data: ",auc_value[1]))


# Plot the ROC curve
pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_ROC_test3.pdf"),height = 2.5, width = 2.5) 
ggroc(roc_obj, legacy.axes = TRUE) +
  geom_text(aes(x = 0.7, y = 0.1, 
                label = paste0("AUC = ", round(auc_value, 2))),
            color = "black") +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black")) +
  xlab("1 - Specificity") + 
  ylab("Sensitivity")
dev.off()


```


# Use marker gene GSVA score to predict ICB response in bulk RNAseq of skin cancer. Conlcusion: not work
```{r}

#### Data (n=121): Liu, D., Schilling, B., Liu, D. et al. Integrative molecular and clinical modeling of clinical outcomes to PD1 blockade in patients with metastatic melanoma. Nat Med 25, 1916–1927 (2019). https://doi.org/10.1038/s41591-019-0654-5

raw_exp_fn = "/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/Liu_melanoma_bulk/41591_2019_654_MOESM3_ESM.txt"
exp_mat = read.csv(raw_exp_fn, sep='\t', header=T, row.names = 1)
pseudobulk_mat = t(scale(exp_mat)) # scale works for each column
print(colnames(pseudobulk_mat))

# extract phenotype for each sample
clinic_info_fn = "/data/Lab_ruppin/tiangen/CancerProject/02.PENCIL/03.Results/Liu_melanoma_bulk/clinicInfo.csv"
clinic_info = read.csv(clinic_info_fn, header = T, row.names = 1)
clinic_info$meta = row.names(clinic_info)
phenotype_df = clinic_info[c("meta", "BR")]
phenotype_df$BR[phenotype_df$BR=="CR"] = 1
phenotype_df$BR[phenotype_df$BR=="PR"] = 1
phenotype_df$BR[phenotype_df$BR=="PD"] = 0
phenotype_df$BR[phenotype_df$BR=="SD"] = 0
phenotype_df$BR[phenotype_df$BR=="MR"] = 0

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# Calculate AUC using pROC package
response <- ifelse(gsva_scores$BR > 0.5, 1, 0)
predictor = as.numeric(gsva_scores$GSVA_diff)
roc_obj <- roc(response, predictor, direction="<")
# Print the AUC value
auc_value <- auc(roc_obj)
print(paste("ICB prediction AUC for test5 data: ",auc_value[1]))


# Plot the ROC curve
pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_ROC_test3.pdf"),height = 2.5, width = 2.5) 
ggroc(roc_obj, legacy.axes = TRUE) +
  geom_text(aes(x = 0.7, y = 0.1, 
                label = paste0("AUC = ", round(auc_value, 2))),
            color = "black") +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black")) +
  xlab("1 - Specificity") + 
  ylab("Sensitivity")
dev.off()


```
