---
title: "Predict ICB response in skin cancer using PENCIL predicted ICBresponse-relevant CD8+T cells"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# load required package
```{r, include=FALSE}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("UCell")
library(UCell)

#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)

library(caret)
library(Seurat)
library(GEOquery)
library(dplyr)
library(ggplot2)
library(SingleR)
library(sctransform)
library(celldex)
library(dittoSeq)
library(SeuratWrappers)
library(harmony)
library(scuttle)
library(scran)
library(scFeatureFilter)
library(cowplot)
library(patchwork)
library(edgeR)
library(RColorBrewer)
library(stringr)
#BiocManager::install("GSVA")
library(GSVA)
library(pROC)

```

# set input and output directories & parameters
```{r}

phenotype_name = 'ICBresponse'
sampleType_name = 'Tissue'
cellType_of_interest = 'CD8T' 

shared_data_dir = paste0("../02.Input/")
PENCIL_result_dir = paste0("./01.Scripts/results/")
PENCIL_result_dir_deep = paste0(PENCIL_result_dir, 'GSE120575_Tissue_CD8T/py/',phenotype_name,'/')

dataset_train = 'GSE120575'
processed_dir_train <- paste0("../03.Results/", dataset_train, '/')
fig_dir_train = paste0("../03.Results/", dataset_train, "/Figures/")

dataset_test1 = 'GSE123813'
processed_dir_test1 <- paste0("../03.Results/", dataset_test1, '/')
fig_dir_test1 = paste0("../03.Results/", dataset_test1, "/Figures/")

dataset_test2 = 'GSE166181'
processed_dir_test2 <- paste0("../03.Results/", dataset_test2, '/')
fig_dir_test2 = paste0("../03.Results/", dataset_test2, "/Figures/")


```


# Identify DEGs for PENCIL predicted N/NR cells on training data
```{r}

logFCfilter = 0.25
adjPvalFilter = 0.05

# loading seurat object of training data
merged_seu = readRDS(file = paste0(processed_dir_train,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))

##### find all markers for PENCIL predicted labels
Idents(object = merged_seu) <- merged_seu@meta.data$'PENCIL_pred_label_ICBresponse'
Cells.markers = FindAllMarkers(merged_seu, only.pos=T, min.pct=0.25, logfc.threshold = logFCfilter)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'

sig.markers = Cells.markers[(abs(as.numeric(as.vector(Cells.markers$avg_log2FC))) > logFCfilter & 
                                 as.numeric(as.vector(Cells.markers$p_val_adj)) < adjPvalFilter), ]

write.csv(sig.markers, paste0(processed_dir_train, "sig.markers_PENCILpredLabel_",sampleType_name, '_', cellType_of_interest,".csv"))

##### Keep only the top N genes for each PENCIL predicted label, ranked by logFC
topN = 5
TopN.markers <- sig.markers %>%
    group_by(cluster) %>%
    top_n(topN, avg_log2FC)

R_markerGenes1 = TopN.markers$gene[TopN.markers$cluster=='yes'] #  "IL7R","CCR7","TSPYL2","GPR183","HSPH1"
NR_markerGenes1 = TopN.markers$gene[TopN.markers$cluster=='no'] # "IFI6","WARS","GZMB","CCL3","ISG15"

```

# plot dotplot for the topN marker genes (Figure 2b)
```{r}

merged_seu_sub = subset(merged_seu, subset = PENCIL_pred_label_ICBresponse != "rej")
merged_seu_sub$PENCIL_pred_label_ICBresponse = as.character(merged_seu_sub$PENCIL_pred_label_ICBresponse)
merged_seu_sub$PENCIL_pred_label_ICBresponse[merged_seu_sub$PENCIL_pred_label_ICBresponse=='no'] = 'Non-responder'
merged_seu_sub$PENCIL_pred_label_ICBresponse[merged_seu_sub$PENCIL_pred_label_ICBresponse=='yes'] = 'Responder'

p <- DotPlot(merged_seu_sub, 
             features = c(R_markerGenes,NR_markerGenes), 
             group.by = "PENCIL_pred_label_ICBresponse",
             cols = c("lightgrey", "red")) + 
  labs(x="", y="")+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 15),
        axis.text.y = element_text(color = "black", size = 15),
        legend.position = c(1.25, 0.3),
        legend.key.height = unit(5, "mm"),
        plot.margin = margin(t = 1, r = 40, b = 1, l = 1, unit = "mm"))
ggsave(filename = paste0(PENCIL_result_dir_deep, "Dotplot_TopNmarkers_train.pdf"), plot = p, height = 30*2.5, width = 40*4, units = "mm")


```


# calculate topN marker genes using CD8T cells from R and NR samples directly (without using PENCIL)
```{r}

##### find all markers for PENCIL predicted labels
Idents(object = merged_seu) <- merged_seu@meta.data$'ICBresponse'
Cells.markers = FindAllMarkers(merged_seu, only.pos=T, min.pct=0.25, logfc.threshold = logFCfilter)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'

sig.markers = Cells.markers[(abs(as.numeric(as.vector(Cells.markers$avg_log2FC))) > logFCfilter & 
                               as.numeric(as.vector(Cells.markers$p_val_adj)) < adjPvalFilter), ]

write.csv(sig.markers, paste0(processed_dir_train, "sig.markers_trueLabel_",sampleType_name, '_', cellType_of_interest,".csv"))

##### Keep only the top N genes for each PENCIL predicted label, ranked by logFC
topN = 5
TopN.markers <- sig.markers %>%
    group_by(cluster) %>%
    top_n(topN, avg_log2FC)

R_markerGenes2 = TopN.markers$gene[TopN.markers$cluster=='1'] # "MT-CYB" "IL7R"   "TSPYL2" "YPEL5"  "RGPD5" 
NR_markerGenes2 = TopN.markers$gene[TopN.markers$cluster=='0'] # "CD38" "IFI6" "GZMB" "CCL3" "WARS"

```


# test predictive power for ICB response of different GSVA signatures (Figure 2cd; Supplmentary Figure 1a-f)
```{r}

marker_gene_source = "PENCIL" # PENCIL  PENCIL_Ren  trueLabel  ExhT   CytT   TIS  IFNr

if (marker_gene_source == "PENCIL"){ ### use marker genes from PENCIL predicted R vs NR
  R_markerGenes = c("IL7R","CCR7","TSPYL2","GPR183","HSPH1")
  NR_markerGenes = c("IFI6","WARS","GZMB","CCL3","ISG15")
}else if (marker_gene_source == "trueLabel"){ ### use marker genes directly from R vs NR
  R_markerGenes = c("MT-CYB","IL7R","TSPYL2","YPEL5","RGPD5")
  NR_markerGenes = c("CD38","IFI6","GZMB","CCL3","WARS")
}else if (marker_gene_source == "PENCIL_Ren"){ ### use marker genes from Ren et al. original PENCIL paper
  R_markerGenes = c("IL7R", "GPR183", "CCR7", "NR4A2", "LMNA", "NR4A3", "CD55", "MGAT4A", "PER1", "TSPYL2", "FOXP1", "STAT4")
  NR_markerGenes = c("CD38", "CCL3", "VCAM1", "HAVCR2", "CCR5", "GOLIM4", "PRDX3", "LGALS9", "ENTPD1", "TRAFD1", "PDCD1", "BATF", "LAG3", "CTLA4", "PTPN6", "CXCR6")
  R_markerGenes = R_markerGenes[1:5]
  NR_markerGenes = NR_markerGenes[1:5]
}else if (marker_gene_source == "ExhT"){  ### use the well-established exhausted T cell signature
  R_markerGenes = c("PDCD1", "CTLA4", "LAG3", "HAVCR2", "TIGIT")
  NR_markerGenes = c("")
}else if (marker_gene_source == "CytT"){  ### use the well-established cytotoxic T cell signature
  R_markerGenes = c("GZMA", "PRF1")
  NR_markerGenes = c("")
}else if (marker_gene_source == "TIS"){  ### use the well-established tumor inflammatory signature
  R_markerGenes = c("CD276", "HLA-DQA1", "CD274", "IDO1", "HLA-DRB1", "HLA-E", "CMKLR1", "PDCD1LG2", "PSMB10", "LAG3", "CXCL9", "STAT1", "CD8A", "CCL5", "NKG7", "TIGIT", "CD27", "CXCR6")
  NR_markerGenes = c("")
}else if (marker_gene_source == "IFNr"){  ### use the well-established IFN-gamma signature
  R_markerGenes = c("IDO1", "CXCL10", "CXCL9", "HLA-DRA", "STAT1", "IFNG")
  NR_markerGenes = c("")
}




min_cellNum_per_sample = 50

gene_set_list <- list(
  GeneSet1 = R_markerGenes,
  GeneSet2 = NR_markerGenes
)

result_list = list() # store GSVA score and ICB response info of different datasets

#### 1) the training set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_train,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
if (is.null(gsva_scores$GeneSet1)){
  gsva_scores$GeneSet1 = 0
}
if (is.null(gsva_scores$GeneSet2)){
  gsva_scores$GeneSet2 = 0
}
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# save to result
response <- ifelse(gsva_scores$ICBresponse > 0, 1, 0) 
predictor = as.numeric(gsva_scores$GSVA_diff)
result_list[['train']] = data.frame(ICBresponse = response, GSVAscore = predictor)



#### 2) the test1 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test1,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))
#merged_seu$meta = merged_seu$patient
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
if (is.null(gsva_scores$GeneSet1)){
  gsva_scores$GeneSet1 = 0
}
if (is.null(gsva_scores$GeneSet2)){
  gsva_scores$GeneSet2 = 0
}
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# save to result
response <- ifelse(gsva_scores$ICBresponse > 0, 1, 0)
predictor = as.numeric(gsva_scores$GSVA_diff)
result_list[['test1']] = data.frame(ICBresponse = response, GSVAscore = predictor)



#### 3) the test2 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test2,"seu_PBMC_CD8T_",phenotype_name,"_PENCIL.rds"))
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
# make pseudo-bulk CD8T expression of each sample
merged_seu$meta <- factor(merged_seu$meta)
Idents(object = merged_seu) <- merged_seu@meta.data$'meta'
pseudobulk_expression <- AverageExpression(merged_seu, slot = "data") # gives normalized value for each sample (10000)
Idents(object = merged_seu) <- merged_seu@meta.data$'seurat_clusters'
pseudobulk_mat_t <- t(pseudobulk_expression$RNA)
pseudobulk_mat = t(scale(pseudobulk_mat_t)) # scale works for each column
print(colnames(pseudobulk_mat))
print(levels(merged_seu@meta.data$'meta'))
colnames(pseudobulk_mat) = levels(merged_seu@meta.data$'meta')

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]
rownames(phenotype_df) = phenotype_df$meta

# Calculate the GSVA scores
gsva_scores <- gsva(pseudobulk_mat, gene_set_list, method = "gsva")
gsva_scores = data.frame(t(gsva_scores))
if (is.null(gsva_scores$GeneSet1)){
  gsva_scores$GeneSet1 = 0
}
if (is.null(gsva_scores$GeneSet2)){
  gsva_scores$GeneSet2 = 0
}
gsva_scores$GSVA_diff = gsva_scores$GeneSet1 - gsva_scores$GeneSet2
gsva_scores$meta = rownames(gsva_scores)
gsva_scores <- merge(gsva_scores, phenotype_df, by = c("meta", "meta"), all = TRUE)

# save to result
response <- ifelse(gsva_scores$ICBresponse > 0, 1, 0)
predictor = as.numeric(gsva_scores$GSVA_diff)
result_list[['test2']] = data.frame(ICBresponse = response, GSVAscore = predictor)

ROC_list <- lapply(result_list, function(df) {
  if (marker_gene_source == "ExhT"){
    roc_curve <- roc(df$ICBresponse, df$GSVAscore, direction = ">") # , levels = c(0, 1), direction = "<"
  }else{
    roc_curve <- roc(df$ICBresponse, df$GSVAscore, direction = "<") # , levels = c(0, 1), direction = "<"
  }
  auc_value <- auc(roc_curve)
  acc_value = mean(df$ICBresponse == (df$GSVAscore>0))
  return(list(roc_curve = roc_curve, auc_value = auc_value, acc_value = acc_value))
})



#### Plot the ROC curves and AUC values
pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_ROC_",marker_gene_source,".pdf"),height = 2.5, width = 2.5) 
ggplot() +
  #theme_minimal() +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black"))+
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(x = "1 - Specificity", y = "Sensitivity") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
  geom_line(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "black", linetype = "solid", linewidth = 1) +
  geom_line(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#FD3A4A", linetype = "solid", linewidth = 1) +
  geom_line(data = data.frame(specificity=ROC_list[[3]]$roc_curve$sensitivities, sensitivity=ROC_list[[3]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#1F8A70", linetype = "solid", linewidth = 1) +
  geom_text(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 0.35, y = 1, label = paste("AUC (train) =", round(ROC_list[[1]]$auc_value, 2))), color = "black",fontface = "plain") +
  geom_text(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 0.35, y = 0.9, label = paste("AUC (test1) =", round(ROC_list[[2]]$auc_value, 2))), color = "#FD3A4A",fontface = "plain") +
  geom_text(data = data.frame(specificity=ROC_list[[3]]$roc_curve$sensitivities, sensitivity=ROC_list[[3]]$roc_curve$specificities), aes(x = 0.35, y = 0.8, label = paste("AUC (test2) =", round(ROC_list[[3]]$auc_value, 2))), color = "#1F8A70",fontface = "plain")#+
dev.off()



### boxplot of GSVA scores in R and NR samples
combined_df <- do.call(rbind, result_list)
pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_boxplot_",marker_gene_source,".pdf"),height = 2.5*0.7, width = 3) 
ggplot() +
  geom_boxplot(data = result_list[[1]], aes(x = factor(ICBresponse, level=c(1,0)), y = GSVAscore, fill = factor(ICBresponse, level=c(1,0)))) +
  geom_boxplot(data = result_list[[2]], aes(x = factor(ICBresponse+2, level=c(3,2)), y = GSVAscore, fill = factor(ICBresponse, level=c(1,0)))) +
  geom_boxplot(data = result_list[[3]], aes(x = factor(ICBresponse+4, level=c(5,4)), y = GSVAscore, fill = factor(ICBresponse, level=c(1,0)))) +
  labs(x = "", y = "Score") +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color="black"),
        axis.text.x = element_text(color="black"),
        legend.position = "none")
dev.off()

print(wilcox.test(result_list[[1]]$GSVAscore[result_list[[1]]$ICBresponse==1], result_list[[1]]$GSVAscore[result_list[[1]]$ICBresponse==0]))
print(wilcox.test(result_list[[2]]$GSVAscore[result_list[[2]]$ICBresponse==1], result_list[[2]]$GSVAscore[result_list[[2]]$ICBresponse==0]))
print(wilcox.test(result_list[[3]]$GSVAscore[result_list[[3]]$ICBresponse==1], result_list[[3]]$GSVAscore[result_list[[3]]$ICBresponse==0]))

```




# test predictive power for ICB response of different CD8+ T cell frequency signatures (Supplmentary Figure 1g-i)
```{r}

marker_gene_source = c("CD39CD103") # CXCL13   CD39   CD39CD103

if (marker_gene_source == "CXCL13"){
  marker_genes = c("CXCL13") 
  accuracy_cutoff = 0.15
}else if (marker_gene_source == "CD39"){
  marker_genes = c("ENTPD1") 
}else if (marker_gene_source == "CD39CD103"){
  marker_genes = c("ENTPD1", "ITGAE") 
}

min_cellNum_per_sample = 50

result_list = list()

#### 1) the training set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_train,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
if (length(marker_genes)==1){
  geneExpr_01 = as.numeric(merged_seu[["RNA"]]@data[marker_genes,]>0)
}else if (length(marker_genes)==2){
  geneExpr_01 = as.numeric(merged_seu[["RNA"]]@data[marker_genes[1],]>0 & merged_seu[["RNA"]]@data[marker_genes[2],]>0)
}
sample_geneExpr = data.frame(sample = merged_seu$meta, geneExpr_01 = geneExpr_01)
sample_geneExpr2 <- sample_geneExpr %>%
  group_by(sample) %>%
  summarize(ratio = mean(geneExpr_01))
sample_geneExpr2 = data.frame(sample_geneExpr2)
colnames(sample_geneExpr2) = c("meta", "freq")

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]

sig_scores <- merge(sample_geneExpr2, phenotype_df, by = c("meta", "meta"), all = T)
# save to result
response <- ifelse(sig_scores$ICBresponse > 0, 1, 0) 
predictor = as.numeric(sig_scores$freq)
result_list[['train']] = data.frame(ICBresponse = response, score = predictor)



#### 2) the test1 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test1,"seu_Tissue_CD8T_",phenotype_name,"_PENCIL.rds"))
#merged_seu$meta = merged_seu$patient
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
if (length(marker_genes)==1){
  geneExpr_01 = as.numeric(merged_seu[["RNA"]]@data[marker_genes,]>0)
}else if (length(marker_genes)==2){
  geneExpr_01 = as.numeric(merged_seu[["RNA"]]@data[marker_genes[1],]>0 & merged_seu[["RNA"]]@data[marker_genes[2],]>0)
}
sample_geneExpr = data.frame(sample = merged_seu$meta, geneExpr_01 = geneExpr_01)
sample_geneExpr2 <- sample_geneExpr %>%
  group_by(sample) %>%
  summarize(ratio = mean(geneExpr_01))
sample_geneExpr2 = data.frame(sample_geneExpr2)
colnames(sample_geneExpr2) = c("meta", "freq")

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]

sig_scores <- merge(sample_geneExpr2, phenotype_df, by = c("meta", "meta"), all = T)
# save to result
response <- ifelse(sig_scores$ICBresponse > 0, 1, 0) 
predictor = as.numeric(sig_scores$freq)
result_list[['test1']] = data.frame(ICBresponse = response, score = predictor)




#### 3) the test2 set (using CD8T pseudo-bulk)
merged_seu = readRDS(file = paste0(processed_dir_test2,"seu_PBMC_CD8T_",phenotype_name,"_PENCIL.rds"))
sample_cellNum = sort(table(merged_seu$meta))
keep_sample = names(sample_cellNum)[sample_cellNum > min_cellNum_per_sample]
merged_seu = subset(merged_seu, subset = meta %in% keep_sample)
if (length(marker_genes)==1){
  geneExpr_01 = as.numeric(merged_seu[["RNA"]]@data[marker_genes,]>0)
}else if (length(marker_genes)==2){
  geneExpr_01 = as.numeric(merged_seu[["RNA"]]@data[marker_genes[1],]>0 & merged_seu[["RNA"]]@data[marker_genes[2],]>0)
}
sample_geneExpr = data.frame(sample = merged_seu$meta, geneExpr_01 = geneExpr_01)
sample_geneExpr2 <- sample_geneExpr %>%
  group_by(sample) %>%
  summarize(ratio = mean(geneExpr_01))
sample_geneExpr2 = data.frame(sample_geneExpr2)
colnames(sample_geneExpr2) = c("meta", "freq")

# extract phenotype for each sample
phenotype_df = merged_seu@meta.data[c('meta', phenotype_name)]
duplicated_rows <- duplicated(phenotype_df[, 1])
phenotype_df <- phenotype_df[!duplicated_rows, ]

sig_scores <- merge(sample_geneExpr2, phenotype_df, by = c("meta", "meta"), all = T)
# save to result
response <- ifelse(sig_scores$ICBresponse > 0, 1, 0) 
predictor = as.numeric(sig_scores$freq)
result_list[['test2']] = data.frame(ICBresponse = response, score = predictor)




ROC_list <- lapply(result_list, function(df) {
  roc_curve <- roc(df$ICBresponse, df$score, direction = "<")
  auc_value <- auc(roc_curve)
  acc_value = mean(df$ICBresponse == (df$score>accuracy_cutoff))
  return(list(roc_curve = roc_curve, auc_value = auc_value, acc_value = acc_value))
})



#### Plot the ROC curves and AUC values
if (marker_gene_source == "CXCL13"){
  pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_ROC_",marker_gene_source,".pdf"),height = 2.5, width = 2.5) 
  ggplot() +
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color="black"),
          axis.line.x = element_line(color="black"),
          axis.ticks.y = element_line(color="black"),
          axis.ticks.x = element_line(color="black"),
          axis.text.y = element_text(color="black"),
          axis.text.x = element_text(color="black"))+
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    geom_line(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "black", linetype = "solid", linewidth = 1) +
    geom_line(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#FD3A4A", linetype = "solid", linewidth = 1) +
    geom_text(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 0.6, y = 0.3, label = paste("AUC (train) =", round(ROC_list[[1]]$auc_value, 2))), color = "black",fontface = "plain") +
    geom_text(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 0.6, y = 0.2, label = paste("AUC (test1) =", round(ROC_list[[2]]$auc_value, 2))), color = "#FD3A4A",fontface = "plain")
  dev.off()
  
  ### boxplot of GSVA scores in R and NR samples
  combined_df <- do.call(rbind, result_list)
  pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_boxplot_",marker_gene_source,".pdf"),height = 2.5*0.7, width = 2.3) 
  ggplot() +
    geom_boxplot(data = result_list[[1]], aes(x = factor(ICBresponse, level=c(1,0)), y = score, fill = factor(ICBresponse, level=c(1,0)))) +
    geom_boxplot(data = result_list[[2]], aes(x = factor(ICBresponse+2, level=c(3,2)), y = score, fill = factor(ICBresponse, level=c(1,0)))) +
    labs(x = "", y = "Score") +
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color="black"),
          axis.line.x = element_line(color="black"),
          axis.ticks.y = element_line(color="black"),
          axis.ticks.x = element_line(color="black"),
          axis.text.y = element_text(color="black"),
          axis.text.x = element_text(color="black"),
          legend.position = "none")
  dev.off()
}else{
  pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_ROC_",marker_gene_source,".pdf"),height = 2.5, width = 2.5) 
  ggplot() +
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color="black"),
          axis.line.x = element_line(color="black"),
          axis.ticks.y = element_line(color="black"),
          axis.ticks.x = element_line(color="black"),
          axis.text.y = element_text(color="black"),
          axis.text.x = element_text(color="black"))+
    coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(x = "1 - Specificity", y = "Sensitivity") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray") +
    geom_line(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "black", linetype = "solid", linewidth = 1) +
    geom_line(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#FD3A4A", linetype = "solid", linewidth = 1) +
    geom_line(data = data.frame(specificity=ROC_list[[3]]$roc_curve$sensitivities, sensitivity=ROC_list[[3]]$roc_curve$specificities), aes(x = 1 - specificity, y = sensitivity), color = "#1F8A70", linetype = "solid", linewidth = 1) +
    geom_text(data = data.frame(specificity=ROC_list[[1]]$roc_curve$sensitivities, sensitivity=ROC_list[[1]]$roc_curve$specificities), aes(x = 0.66, y = 0.23, label = paste("AUC (train) =", round(ROC_list[[1]]$auc_value, 2))), color = "black",fontface = "plain") +
    geom_text(data = data.frame(specificity=ROC_list[[2]]$roc_curve$sensitivities, sensitivity=ROC_list[[2]]$roc_curve$specificities), aes(x = 0.66, y = 0.13, label = paste("AUC (test1) =", round(ROC_list[[2]]$auc_value, 2))), color = "#FD3A4A",fontface = "plain") +
    geom_text(data = data.frame(specificity=ROC_list[[3]]$roc_curve$sensitivities, sensitivity=ROC_list[[3]]$roc_curve$specificities), aes(x = 0.66, y = 0.03, label = paste("AUC (test2) =", round(ROC_list[[3]]$auc_value, 2))), color = "#1F8A70",fontface = "plain")#+
  dev.off()
  
  ### boxplot of GSVA scores in R and NR samples
  combined_df <- do.call(rbind, result_list)
  pdf(file = paste0(PENCIL_result_dir_deep, "GSVA_sig_boxplot_",marker_gene_source,".pdf"),height = 2.5*0.7, width = 3) 
  ggplot() +
    geom_boxplot(data = result_list[[1]], aes(x = factor(ICBresponse, level=c(1,0)), y = score, fill = factor(ICBresponse, level=c(1,0)))) +
    geom_boxplot(data = result_list[[2]], aes(x = factor(ICBresponse+2, level=c(3,2)), y = score, fill = factor(ICBresponse, level=c(1,0)))) +
    geom_boxplot(data = result_list[[3]], aes(x = factor(ICBresponse+4, level=c(5,4)), y = score, fill = factor(ICBresponse, level=c(1,0)))) +
    labs(x = "", y = "Score") +
    theme(panel.grid = element_blank(), 
          panel.background = element_rect(fill = "white"),
          axis.line.y = element_line(color="black"),
          axis.line.x = element_line(color="black"),
          axis.ticks.y = element_line(color="black"),
          axis.ticks.x = element_line(color="black"),
          axis.text.y = element_text(color="black"),
          axis.text.x = element_text(color="black"),
          legend.position = "none")
  dev.off()
}


print(wilcox.test(result_list[[1]]$score[result_list[[1]]$ICBresponse==1], result_list[[1]]$score[result_list[[1]]$ICBresponse==0]))
print(wilcox.test(result_list[[2]]$score[result_list[[2]]$ICBresponse==1], result_list[[2]]$score[result_list[[2]]$ICBresponse==0]))
print(wilcox.test(result_list[[3]]$score[result_list[[3]]$ICBresponse==1], result_list[[3]]$score[result_list[[3]]$ICBresponse==0]))

```
